HI-FISH: Whole brain in situ mapping of neuronal activation in Drosophila during social behaviors and optogenetic stimulation

Version 1

Abstract

Abstract

Monitoring neuronal activity at single-cell resolution in freely moving Drosophila engaged in social behaviors is challenging because of their small size and lack of transparency. Extant methods, such as Flyception, are highly invasive. Whole-brain calcium imaging in head-fixed, walking flies is feasible but the animals cannot perform the consummatory phases of social behaviors like aggression or mating under these conditions. This has left open the fundamental question of whether neurons identified as functionally important for such behaviors using loss-or gain-of-function screens are actually active during the natural performance of such behaviors, and if so during which phase(s). Here we describe a method, called HI-FISH, for brain-wide mapping of active cells expressing the Immediate Early Gene hr38 using a high-sensitivity/low background amplification method called HCR-3.0. Using double-labeling for hr38 mRNA and for GFP, we describe the activity of several classes of aggression-promoting neurons during courtship and aggression, including P1a cells, an intensively studied population of male-specific interneurons. Using HI-FISH in combination with optogenetic activation of aggression-promoting neurons (opto-HI-FISH) we identify candidate downstream functional targets of these cells in a brain-wide, unbiased manner. Finally we compare the activity of P1a neurons during sequential performance of courtship and aggression, using intronic vs. exonic hr38 probes to differentiate newly synthesized nuclear transcripts from cytoplasmic transcripts synthesized at an earlier time. These data provide evidence suggesting that different subsets of P1a neurons may be active during courtship vs. aggression. HI-FISH and associated methods may help to fill an important lacuna in the armamentarium of tools for neural circuit analysis in Drosophila.

Introduction

Monitoring neuronal activity during specific actions or activities is critical to understanding how the brain controls behavior. Measurements within a local brain region of neural activity in non-transparent awake, behaving animals such as mice or fruit flies can be achieved with electrophysiological and calcium imaging techniques (Buzsáki, 2004; Grewe and Helmchen, 2009). In Drosophila, two-photon calcium imaging (TPI) has been widely used (Seelig et al., 2010) because single-unit electrical recording in the fly central brain remains exceedingly difficult (Turner et al., 2008; Wilson et al., 2004). But TPI in Drosophila requires head-fixation, which limits the repertoire of behaviors that flies can perform under these conditions. For example, it is difficult to use TPI to image neuronal activity during the consummatory phases of complex social behaviors, such as male-male aggression or male-female courtship (but see Clowney et al., 2015). Recently, a new method called Flyception2 was introduced for brain imaging in freely walking flies (Grover et al., 2020, 2016). However the spatial resolution of this method is limited, it is invasive, and monitoring the activity of single neurons in freely moving animals is still challenging.

In mice, immediate-early genes (IEGs) such as c-fos, Arc, or Egr1 have been used to map the activity of neurons by visualizing their expression histologically using immunohistochemistry (IHC) or in situ hybridization (ISH) (Morgan and Curran, 1991). More recently, this method has been combined with tissue-clearing and light-sheet microscopy to perform systematic and unbiased whole-brain mapping of neurons activated during different behaviors (Kim et al., 2015; Renier et al., 2016). An extension of this approach, called cellular compartment analysis of temporal activity by fluorescence in situ hybridization (catFISH), enables within-animal comparisons of neuronal populations activated during two different sequential behaviors (Guzowski et al., 1999; Lin et al., 2011).

In Drosophila, by contrast, the application of IEGs to map whole-brain neuronal activation patterns has been limited. In part this is because c-fos is not a particularly good activity marker in flies (Chen et al., 2016). More recently, several studies using fly IEGs such as Hr38 have been reported (Chen et al., 2016; Fujita et al., 2013; Takayanagi-Kiya and Kiya, 2019). Nevertheless, this technique has been relatively under-utilitized in flies. This may be the case, at least in part, because application of FISH to the adult Drosophila brain has been technically challenging, due to limited probe penetration and high background caused by traditional amplification methods such as Tyramide (Raap et al., 1995).

To circumvent these problems, we have applied in this study an in situ hybridization amplification method called the Hybridization Chain Reaction (HCR), v.3.0 (Choi et al., 2018) to visualize the expression of Hr38, one of the few IEGs shown to work in the adult male fly brain (Fujita et al., 2013). The design of HCR, which requires conjoint hybridization to target sequences of adjacent pairs of “half-probes” to achieve amplification, affords both greater sensitivity and specificity, allowing detection of low-abundance transcripts with minimal background noise and off-target hybridization. We call this method HCR3.0-amplified IEG Fluorescent In Situ Hybridization, or “HI-FISH” (Figure 1A).

Mapping neuronal activities using Hr38 FISH.

(A) Illustrated summary of this study for mapping Hr38 expression induced by social behavior or artificial activation of specific neuronal populations. (B) Representative images of Hr38 expression before (fru-No PS) and after photo-stimulation of fru-GAL4>Chrimson neurons (fru-PS). BDPG4uw-PS represents the images after Photo-stimulation of empty-GAL4>Chrimson neurons. magenta: Chrimson::tdT, green: Hr38 HCR signals. (C) High magnification images of Hr38 expression after photo-stimulation of fru-GAL4>Chrimson neurons. SMP, superior medial protocerebrum; LH, lateral horn; KC, Kenyon cells.

Here we provide proof-of-concept for three different applications of HI-FISH relevant to understanding the neural circuitry underlying social behaviors in flies. First, we have asked whether specific neuronal populations discovered in functional screens for aggression-promoting neurons (Asahina et al., 2014; Hoopfer et al., 2015; Watanabe et al., 2017) are indeed active during natural aggressive behavior, courtship or both. We have further investigated whether such activity is only observed during the contact-mediated consummatory phase of these behaviors, or can also be detected in freely moving flies in the absence of fighting or mating. Second, we have performed unbiased, whole-brain functional mapping of the downstream targets activated by optogenetic stimulation of a neuronal population of interest, a variant method we call “opto-HI-FISH.”. We have applied this method to P1a neurons (Anderson, 2016; Clowney et al., 2015; Hoopfer et al., 2015; Inagaki et al., 2014) a subset of the male-specific P1 class of interneurons originally identified based on their ability to trigger male courtship behavior when stimulated (von Philipsborn et al., 2011). Finally, we have developed a variant of the catFISH technique (Guzowski et al., 1999), using dual-color FISH combining intronic and exonic Hr38 probes, to compare neuronal populations activated during two different sequential behaviors, a method we call “HI-catFISH.” Since experimental activation of P1a neurons can promote both aggression and courtship in pairs of male flies (Hoopfer et al., 2015), we have used HI-catFISH to investigate whether separate or overlapping subpopulations of P1a neurons are active during natural occurrences of these two social behaviors. Together, the results of these studies illustrate the utility of HI-FISH, opto-HI-FISH and HI-catFISH to detect the activation of both known and previously unidentified neurons during naturalistic social behaviors that are not amenable to head-fixation. The results have also yielded new biological insights into the neural circuit-level control of aggression and courtship.

Results

Brain-wide expression of Hr38 detected by HI-FISH

The HCR3.0 method provides greater sensitivity, specificity and lower background than traditional FISH using Tyramide signal amplification (TSA). Application of older FISH methods in the Drosophila adult whole brain has been limited, due to low sensitivity and high background (Wilkie and Davis, 1998). For example, the detection by FISH of low-abundance transcripts such as those encoding GPCRs has been virtually impossible in adult flies up to now. Application of the HCR-FISH method, by contrast, was able to detect mRNA encoding an octopamine receptor Oct-TyrR (Arakawa et al., 1990; Saudou et al., 1990), as well as Dh44, a high-abundance neuropeptide-encoding transcript (Figure 1-figure supplement A and B) (Choi et al., 2018). Background in controls lacking the HCR initiator probes was essentially undetectable.

Next, we investigated the ability of HCR to detect brain-wide expression of Hr38 following artificial (optogenetic) activation of fruitless-GAL4 (Stockinger et al., 2005) neurons expressing the red-shifted opsin Chrimson::tdT. Fru-GAL4 labels ∼2,000 neurons in the adult male brain, providing a large number of potentially activatable cells. Following photostimulation Hr38 signals were detected in virtually all Chrimson::tdT+ neurons as well as in other tdT- neurons (Figure 1B). The latter may be indirectly activated via synaptic inputs from Fru-GAL4+ neurons. Control flies of the same genotype but without photostimulation, or photostimulated flies with an “empty” GAL4 driver (BDPG4uw; Pfeiffer et al., 2008) displayed much lower levels of Hr38 expression.

While performing these experiments, we observed that simply transferring flies to a new arena induced a significant level of Hr38 expression throughout the brain (Figure 1-figure supplement C and D), likely reflecting the stress caused by the manipulation and new environment. Based on this observation, we conducted subsequent experiments after flies were habituated to the testing arena before starting the assays (see Methods). This minor modification proved essential to distinguishing behavior-specific HI-FISH signals from the widespread Hr38 induction caused by transfer of the flies from their home vial to an experimental apparatus.

Mapping of neurons activated by social behaviors

As an initial proof-of-concept application of HI-FISH, we asked whether neuronal subsets initially identified in functional screens for aggression-promoting neurons (Asahina et al., 2014; Hoopfer et al., 2015; Watanabe et al., 2017) were actually active during natural aggressive behavior. These included P1a, Tachykinin-FruM+ (TkFruM) and aSP2 neurons. Despite the fact that activity in these neuron subsets was shown to be essential for normal aggression, it was formally possible that the behavioral phenotypes caused by their artificial stimulation were due to their supra-physiological activation. Therefore, we performed HI-FISH on fly lines carrying a myr:GFP reporter in the cell populations of interest in males paired with different types of target flies under different conditions. Thirty to 45 minutes after a 20 min interaction with the target fly, we fixed the brains and performed double-label HI-FISH together with fluorescent antibody staining of myr::GFP-expressing neurons (Figure 1A; see Methods).

In the control condition, where the experimental flies were kept alone during the observation period, there was little or no detectable Hr38 expression in P1a, Tk, or aSP2 neurons (Figure 2A-D, “Control”). In contrast, we observed clear Hr38 signals in GFP+ P1a, Tk, and aSP2 neurons after a male-male interaction that included fighting behavior (Aggression: Figure 2A-D, Figure 2-figure supplement). In the case of courtship with a female, comparably strong Hr38 induction was observed in P1a and aSP2 neurons, while there was no Hr38 expression in Tk neurons (Figure 2A-D, Figure 2-figure supplement; Courtship). These data are consistent with prior observations indicating that the functional manipulation of P1a or aSP2 neurons in the male flies affects both male-male aggression and male-female courtship behavior (Hoopfer et al., 2015; Takayanagi-Kiya and Kiya, 2019; Watanabe et al., 2017), while the manipulation of Tk neurons affects only aggression (Asahina et al., 2014).

Investigation of neurons activated by social behavior with Hr38 FISH.

(A) Hr38 expression in the neurons labeled with P1a-split GAL4, Tk-GAL4, and aSP2-split GAL4 drivers. magenta: myr::GFP, yellow: Hr38 HCR signals. Dotted lines depicted the outlines of myr::GFP labeled cell bodies. Arrows: cell bodies of Hr38 positive neurons. (B) - (D) Expression level of Hr38 HCR signals (top) and percentage of Hr38 positive neurons (bottom) in P1a (B), Tk (C), and aSP2 (D) after different behavioral episodes. Aggression: aggressive behavior against another male, Dead: interaction with a dead male, No Food: interaction with another male without food (no fighting), Mesh: interaction with another male, separated with mesh (no physical contact), Courtship: courtship behavior with a virgin female. Bottom: They are grouped in categories according to the characters (a, b, c). In each group, the behavioral conditions do not have a statistically significant difference (p < 0.05, chi-square test). Number of neurons analyzed; P1a-Control: 135, P1a-Aggression: 84, P1a-Dead Male: 126, P1a-No Food: 115, P1a-Mesh: 87, P1a-Courtship: 114, Tk-Control: 45, Tk-Aggression: 69, Tk-Dead Male: 61, Tk-No Food: 63, Tk-Mesh: 82, Tk-Courtship: 66, aSP2-Control: 20, aSP2-Aggression: 36, aSP2-Dead Male: 23, aSP2-No Food: 49, aSP2-Mesh: 36, aSP2-Courtship: 24.

Social interactions in flies consist of multiple phases: approach, investigation, initial contact and progression to consummatory behaviors (Chiu et al., 2021; Yamamoto and Koganezawa, 2013). While electrophysiological recording or calcium imaging can identify neuronal activation that is time-locked to specific behavioral actions, the detection of Hr38 (or any IEG) induction may reflect activation at any or all of these sequential phases due to the long-lived, accumulated expression of the IEG mRNA. To identify the phase(s) of social behavior wherein Hr38 induction occurred, we separated the signals or interactions characteristic of these sequential phases in space rather than time. First, we asked whether the activation of these neurons during a male-male social interaction requires actual fighting behavior, or simply close contact with the target male. Initially, we paired the experimental male fly with a dead male, as a source of chemosensory cues. Because some male-derived cues (e.g., visual motion) may require a live conspecific, we also placed the experimental and tester males in an arena with an agarose substrate. Under these conditions, physical contact but not fighting occurs (Lim et al., 2014). In both “touching but no fighting” conditions, the fraction of Hr38+ cells among P1a, Tk and aSP2 neurons was slightly lower but not statistically significantly different from that measured during actual aggression, while the intensity of Hr38 mRNA expression in positive cells was slightly lower in the No Food condition (Dead Male and No Food: Figure 2B-D, Figure 2-figure supplement). Thus the activation of P1a, Tk, or aSP2 neurons does not require consummatory aggressive actions such as lunging, tussling or boxing (Chen et al., 2002).

Interacting male flies can, in principle, exchange visual, auditory, mechanosensory or chemosensory cues. In the case of chemosensory signals, inter-male aggression has been shown to require both detection of volatile cues by the olfactory system, and of non-volatile cues by the gustatory system (Fernández and Kravitz, 2014; Wang et al., 2011; Wang and Anderson, 2010). While both types of cues are present when an experimental fly is exposed to an intact or a dead male fly, the detection of non-volatile cues requires contact chemosensation (Vosshall and Stocker, 2007). We therefore exposed the experimental males to a target fly separated by a mesh filter which prevented physical contact between the freely moving flies. Interestingly, Hr38 induction in all 3 cell populations under this condition was significantly weaker that that measured during free interaction with a male on agarose, where contact but no fighting occurred (Mesh: Figure 2B-D, Figure 2-figure supplement). Taken together with the relatively strong Hr38 induction observed following contact with a dead male (which presumably provide a source of both olfactory and gustatory pheromones), these data suggest that strong activation of P1a, Tk, and aSP2 neurons requires the integration of both volatile and non-volatile male-derived chemosensory cues. However the results do not exclude the possibility that these neurons exhibit additional activity during contact-mediated aggression, but which is not statistically detectable by this method in comparison to “no fighting” controls.

Detection of Hr38+ cells induced by optogenetic stimulation of P1a neurons in the male brain

Optogenetic and thermogenetic stimulation experiments have shown that that P1a interneurons can promote both male-directed aggression and male-or female-directed courtship (Hoopfer et al., 2015; von Philipsborn et al., 2011). While the neural circuitry downstream of P1 neurons that promotes female-directed courtship behavior has been well-characterized (von Philipsborn et al., 2011), the pathway by which P1a neurons can promote aggression is poorly understood. As a first step to identify systematically P1a functional downstream targets, we optogenetically activated P1a neurons using Chrimson:tdT and mapped the brain-wide pattern of Hr38 induction in males (Figure 3A-D), a method we call “opto-HI-FISH (Figure 1A).” To analyze these data, images of Hr38 expression from multiple individuals were registered to a reference brain using antibody labeling of Bruchpilot (Brp), a reference neuropil marker, to generate an average induction pattern (Figure 3E) (Cachero et al., 2010). We then created a voxel-based heat map for the Hr38 expression from the average image. The heat map indicated that neurons around the posterior dorsal region of the central brain were strongly activated by P1a stimulation (Figure 3F).

Investigation of downstream neurons activated by optogenetic activation of P1a neurons.

(A) Illustration of the experimental set up. (B) Expression pattern of Chrimson::tdT driven by P1a split GAL4. (C) Scheme of the data analysis procedure. (D) Time-course study of Hr38 HCR expression after P1a photo-stimulation in P1a labeled neurons and the surrounding area. The expression level reaches its peak at 30-60 mins. (E) Average image of Hr38 expression of control (BDPG4Uw) and P1a-activated brains. 5 individual brains from each conditions were registered into a template brain and averaged. Green: Hr38 HCR signals, Magenta: Brp. (F) Heat map analysis of Hr38 expression. Areas depicted by dotted lines represent the areas of PAM neusons (PAM), mushroom bodies (MB), and pC neurons (pC). (G) Confocal images of Hr38 expressions in pCd neurons (R41A01), Kenyon cells (OK107), and PAM neurons (0273 or Ddc) of control and P1a activated brains. Green: myr::GFP, Magenta: Hr38. Dotted lines depicted the outlines of cell bodies (pCd) or cluster of neurons (KC, PAM).

To validate this approach, we first asked whether we could detect Hr38 induction in pCd neurons, which were previously shown by calcium imaging to be (indirect) targets of P1a neurons. Double-labeling in pCd-GAL4; UAS-GFP flies using HI-FISH and immunostaining for GFP (see Methods) confirmed that Hr38 was strongly induced in pCd neurons by P1a photostimulation (Figure 3G, pCd). Next, we asked whether we could identify any of the specific cellular targets of P1a neurons in brain regions that exhibited strong induction of Hr38. These regions included Kenyon cells (KCs) and PAM dopaminergic neurons in the mushroom body (MB; Figure 3F). We used the GAL4 drivers OK107 (Connolly et al., 1996) to label KCs, and 0273 and Ddc-GAL4 to label PAM neurons. Following P1a optogenetic activation, brains were processed for double-labeling with HI-FISH and anti-GFP antibody staining to identify the cells of interest. We observed evidence of Hr38 induction in all three cell populations, although the signal was difficult to quantify due to the high density of GFP+ cells (Figure 3G).

To confirm the results of the opto-HI-FISH experiments, we next asked whether PAM neurons and/or Kenyon cells were activated in response to P1a stimulation using in vivo calcium imaging in living, head-fixed flies. We expressed a calcium indicator, jGCaMP7b (Dana et al., 2019) in either Kenyon cells or PAM neurons using the GAL4/UAS system and Chrimson::tdT in P1a neurons using the LexA/LexAop system (Brand and Perrimon, 1993; Lai and Lee, 2006). The neurites of both KCs and PAM neurons showed significant time-locked responses to optogenetic activation of P1a neurons (Figure3-figure supplement 3A-F). We also imaged jGCaMP7b signals from cell bodies of PAM neurons while photo-stimulating P1a neurons, and observed clear responses in both cases as well (Figure 3-figure supplement 3G-J). Although we could not identify the specific activated cell populations among the KCs (data not shown), P1a photo-stimulation evoked a significant rise in ΔF/F in ∼50 % of PAM neurons (Figure 3-figure supplement 3I). These data validate the results from brain-wide Hr38 induction mapping and provide proof-of-concept that brain regions containing functional targets of a cell population of interest (here P1a neurons), as well as specific neuron subtype targets within these regions (KCs and PAM neurons), can be identified by opto-HI-FISH. These results are also consistent with a recent report showing that subsets of PAM dopaminergic neurons receive excitatory inputs from P1 neurons, and assign to KCs a positive valence associated with P1 activation (Shen et al., 2023)

HI-catFISH reveals distinct fighting-and mating-activated P1a neuron subsets

Previous reports showed that artificial activation of P1a neurons promoted both aggression and courtship. The foregoing data indicated that Hr38 was induced in P1a neurons after both natural aggressive and courtship behavior (Figure 2, Figure 2-figure supplement). An important outstanding question is whether the aggression-and courtship-activated P1a populations are the same or different (Anderson, 2016). Because individual P1 neuron subtypes cannot be distinguished by FISH, between-animal comparisons of Hr38 induction following fighting or mating cannot distinguish these two possibilities. To compare directly which P1a neurons are active during naturalistic mating vs. aggression, we developed an Hr38 analog of catFISH (HI-catFISH) to compare neurons activated during aggression and courtship in the same animal (Figure 4A). The catFISH method exploits the different kinetics of nuclear-cytoplasmic transport and of IEG mRNA stability (Guzowski et al., 1999) to distinguish neurons activated during a behavior performed just prior to analysis (nuclear pre-mRNA), from those activated during a different behavior performed 30 min before the second (cytoplasmic mRNA); cells activated during both behaviors contain both cytoplasmic and nuclear signals (Figure 4B). In mice, we have used catFISH with c-fos to compare hypothalamic neurons activated during aggression vs. mating in males (Lin et al., 2011).

Drosophila catFISH reveals P1a subpopulations activated by two different social behaviors.

(A) Experimental design of the fly catFISH. Each fly experienced the 1st behavior (aggression or courtship) followed by 2nd behavior (aggression or courtship), and the HCR was performed with the Hr38 probes targeted for the exon or the intron sequence. (B) Time-course of Hr38 signals detected by the exon-or the intron-targeted Hr38 HCR probes and expected Hr38 signals in each neurons. (Top) The intron-targeted Hr38 HCR signals (magenta) induced by 1st behavior disappear at the timing of HCR analysis while the exon-targeted Hr38 signals (blue) remain in the cytoplasmic area. (Middle) Both intron-targeted and exon-targeted signals are present in the nucleus. (Bottom) Both intron-targeted and exon-targeted signals induced by the 2nd behavior are present in the nucleus, and exon-targeted Hr38 signals induced by the 1st behavior are present in the cytoplasmic area. In ’Neurons active during both’, the exon probe signals from the 1st behavior decline (dotted blue line *1) after 60 mins (also see Fig. 4B-supplement). However, it is canceled out as the signals from 2nd behavior are increased (dotted blue line *2). As a result, the combined Hr38 exon probe signals (solid blue line) remain high at the time point for Hr38 HCR. (C) Representative images of Hr38 HCR signals in P1a neurons after four different behavioral conditions. Green: myr::GFP, Magenta: intron-targeted Hr38 HCR signals: exon-targeted Hr38 signals. (D) Fraction of responsive P1a neurons across indicated conditions. N.R.: not responded, Nuc: nuclear signal only (responded to only second target), Cyto: cytoplasmic signal only (responded to only first target), Both: nuclear and cytoplasmic signal (responded to both first and second target).

We detected neuronal populations activated during two consecutive social behaviors using HI-FISH containing Hr38 intronic and exonic probes amplified using different fluorophores. In two color FISH, unspliced pre-mRNAs were detected as nuclear dots labeled by both probes, while mRNA was detected as a more diffuse cytoplasmic signal detected by only the exonic probe (Figure 4B). A time course study using opto-HI-FISH with these two different Hr38 probes indicated that the primary RNA transcripts first appeared in P1a nuclei <15 min after optogenetic activation, and disappeared by 60 min. Conversely, cytoplasmic processed mRNAs started to accumulate in the cytoplasm 30 min after activation and disappeared by 2-3 hours after stimulation (Figure 4B, top two panels and Figure 4-figure supplement 4C).

Having established this calibration of the temporal resolution of HI-catFISH, we compared labeling during free social interactions with a male or a female target. Following pre-habituation (see above), a naïve experimental male fly was placed into a behavioral arena, and a male or female fly was introduced as a first target. After 20 minutes of free social interactions the target fly was removed from the arena and, after a further 30 min recovery period, replaced by a second target fly (male or female; Methods). All four combinations of sex of the 1st and 2nd target fly were tested (Fig. 4C). Ten minutes after exposure to the second target fly (and ∼1 hr after exposure to the first target fly), the experimental fly was sacrificed and analyzed by dual-color HI-FISH Hr38 (Figure 4A and B).

As expected, many of the neurons activated by the first target were re-activated by the second target (nuclear and cytoplasmic signal), when the sex of the two target flies was the same (Figure 4C, Male -> Male, Female -> Female, arrowheads and Figure 4D, “Both” vs. “Cyto”). By contrast, many neurons activated by the first target fly (cytoplasmic signal) were not activated by the second target fly (nuclear signal), when the target sexes were different (Figure 4C, Female -> Male, Male -> Female, arrows and Figure 4D, “Cyto” vs. “Both”). Responses to males were clearly detected in a subset of P1a neurons whether the male was presented first (Figure 4C, MaleèFemale; GFP+; dashed oval outlines containing blue diffuse cytoplasmic signal) or second (Figure 4C, FemaleèMale; magenta nuclear dots). In addition, there were male-activated cells nearby that were not labeled by the P1a-split GAL4 driver. In general, these male HI-FISH signals did not overlap with the HI-FISH signals evoked by female targets. (We note, however, that relatively few neurons were activated by the female fly when the sex of the two targets was different, regardless of whether the female was presented first or second; the reasons for this are currently unclear but may reflect a suppressive effect of aggression neuron activation on courtship neuron activity, as suggested by earlier studies (Hoopfer et al., 2015).) These data suggest that there may be distinct P1a subpopulations that are activated during either male-male aggression or male-female courtship.

Discussion

Drosophila is a powerful model system for identifying and characterizing neural circuits that control innate behaviors. With recent progress in two-photon imaging techniques using genetically encoded calcium sensors, it has become possible to monitor activity from genetically labeled neurons at single-cell resolution in head-fixed flies (Seelig et al., 2010). Although single-neuron imaging in unrestrained behaving animals has been achieved in mammalian studies (Flusberg et al., 2008; Remedios et al., 2017; Ziv et al., 2013) most imaging studies with single-cell resolution in Drosophila require head-fixed conditions. Therefore, the repertoire of naturalistic behaviors that can studied using this approach – in particular social behaviors dependent on chemosensory cues (but see Hindmarsh Sten et al., 2021)– remains limited.

In mammalian systems, the discovery that IEGs such as c-fos are induced by neuronal activity (Greenberg et al., 1985; Greenberg and Ziff, 1984) has enabled brain-wide functional mapping of neuronal activation following specific behaviors (Morgan and Curran, 1991; Renier et al., 2016). This method and its derivatives have had a transformative impact on the identification of behaviorally relevant neural circuits in rodents. In Drosophila, by contrast, few if any studies describe the application of IEGs for brain-wide activity mapping. In this report, we applied the HCR3.0 RNA-FISH technique (Choi et al., 2018) to visualize the expression of a fly IEG, Hr38 in the entire adult male brain following different manipulations (see also Shao et al., 2019). We present data providing proof-of-concept that HI-FISH can be used to detect the activation by naturalistic stimuli of neurons previously identified in thermogenetic behavioral activation screens, as well as to identify candidate downstream targets of these neurons following their optogenetic activation (opto-HI-FISH). We also compared neuronal subsets within a small identified population (P1a neurons) that is activated during two different social behaviors, using HI-catFISH.

Monitoring Neuronal Activities during Social Behaviors

Previous studies reported that male-specific P1a, Tk, and aSP2 neurons promote male-male aggression (Asahina et al., 2014; Hoopfer et al., 2015; Watanabe et al., 2017). In addition, P1a and aSP2 neurons promote male-female courtship (Hoopfer et al., 2015; Takayanagi-Kiya and Kiya, 2019). Artificial manipulation of the activity of these neurons affected social behaviors and demonstrated their causal roles in these behaviors. However, there was no direct evidence showing that these neurons are normally activated during the specific behaviors they can promote.

Our HI-FISH data demonstrated that P1a and aSP2 neurons are active during male-male aggression and male-female courtship, consistent with the behavioral effects of activating these neurons (Figure 2) (Hoopfer et al., 2015; Watanabe et al., 2017). In contrast, TkFruM neurons were only active during male-male aggression (Figure 2), consistent with their lack of a functional role in courtship (Asahina et al., 2014). Furthermore, we show that P1a, Tk, and aSP2 neurons can apparently be activated by chemosensory cues from an opponent male fly in the absence of attack behavior (Figure 2). These data suggest that P1a, Tk, and aSP2 neurons may serve to process and integrate sensory cues from the opponent and relay that information to the downstream neurons that encode the consummatory phase of aggression (Chiu et al., 2021). However they do not exclude that these neurons may play a role during overt attack.

Although the time resolution of IEG-based activity monitoring falls far short of that of calcium imaging, our data show that one can nevertheless use HI-FISH to separate activation that occurs during the appetitive vs. consummatory phases of male-male aggression, by judicious manipulation of experimental conditions.

Identification of downstream target neurons of genetically labeled neurons by opto-HI-FISH

Artificial activation of P1a neurons induces male-female courtship and an internal state that promotes male-male aggression (Bath et al., 2014; Hoopfer et al., 2015; Inagaki et al., 2014). Furthermore, artificial activation of male specific P1 neurons mimics courtship-induced place preference as well as supporting appetitive olfactory conditioning (Shen et al., 2023). It is critical to identify neurons that are functionally downstream of P1a cells to understand how these neurons can promote this state. In this study, using opto-HI-FISH we identified pCd neurons, Kenyon cells and PAM neurons as downstream targets of P1a neurons. These results were confirmed by two-photon calcium imaging analysis (Figure 3, Figure 3-figure supplement). Importantly, recent studies have reached the same conclusions regarding neuronal populations using different approaches (Jung et al., 2020; Shen et al., 2023). These findings provide additional support for our methodology. The heat map illustrates the presence of additional potential P1a -follower populations beyond the three interrogated here (Figure 3F). Further analysis will unveil the detailed characteristics of these potential downstream neurons of P1a neurons. An important caveat is that not all active Drosophila neurons may express Hr38. Act-seq data from mice indicates that different neuron subsets activated during social behaviors express different IEGs to different extents (Kim et al., 2019).

Functional heterogeneity of P1a populations during Social Behaviors

In addition to identifying functional downstream targets of P1a neurons, characterizing cellular diversity among P1a neurons and how this functional heterogeneity maps onto their connectivity with different downstream targets is critical to understanding how these cells can induce both courtship and aggressive behavior. Optogenetic activation of P1a neurons at a low frequency promotes aggression towards males, while high-frequency stimulation promotes male-directed courtship during stimulation, followed by aggression in the inter-stimulation intervals (Hoopfer et al., 2015). P1a neurons may control these two behaviors according to their activity level and/or the subtype of neuron activated in the population. Alternatively, P1a neurons may indirectly promote fighting by inhibiting aggression-promoting neurons during courtship (Koganezawa et al., 2016); post-inhibitory rebound in such neurons following the offset of P1a optogenetic stimulation may trigger aggressive behavior (Anderson, 2016).

Our preliminary Hr38 catFISH results suggest the presence of male-or female-responsive subsets within the P1a population, which consists of 8-10 neurons/hemibrain (Figure 4). To our knowledge, this is the first evidence that P1a neurons are activated during male-male social interactions. Previous studies using calcium imaging of head-fixed flies suggested that P1a neurons were inhibited by the male-specific pheromone, 11-cis-vaccenyl acetate (cVA; Clowney et al., 2015). The reason for this apparent discrepancy is not yet clear. The molecular identity of the male-vs. female-responsive P1a subsets remains to be determined. It has been suggested that FruM+ P1 neurons exclusively control courtship, while Dsx+ P1 neurons (also called pC1 neurons; Koganezawa et al., 2016)) control aggression. However triple-intersection experiments have demonstrated that thermogenetic (Hoopfer et al., 2015) or optogenetic (Watanabe et al., 2017) activation of FruM+ P1a neurons, which comprise only 1-4 neurons/hemibrain, can also promote aggression. Therefore, while our results support the idea that different P1 subsets control aggression vs. courtship, they do not support the view that FruM+ P1 neurons only control the latter and not the former social behavior.

In mice, distinct subpopulations of estrogen receptor 1 (Esr1)-expressing neurons in the ventrolateral subdivision of the ventromedial hypothalamus (VMHvl) are activated during male-directed aggression versus female-directed mating (Karigo et al., 2021; Remedios et al., 2017). Our observation that apparently distinct subpopulations of P1a neurons are activated during aggression versus courtship in Drosophila strengthens the analogy between these two control nodes for social behavior (Anderson, 2016). Application to P1a neurons of Act-Seq, a method that uses scRNAseq to identify transcriptomic cell types that are active during different behaviors as determined by IEG expression (Kim et al., 2019; Wu et al., 2017), may help to identify transcriptomic and functional diversity within the P1a population. Such correlative studies may be extended to functional perturbations by using transcriptomic data to identify more specific intersectional GAL4 drivers for distinct P1a subsets. Together, our studies illustrate how HI-FISH can provide a systematic, unbiased and brain-wide approach to identifying entry points to circuits controlling specific social behaviors, which can be integrated eventually with imaging, connectomic, transcriptomic and functional approaches to circuit analysis.

Note

While this paper was in the final stages of preparation, a manuscript by Takayanagi-Kiya et al. appeared that described a method for labeling neurons activated during innate behaviors using a different immediate early gene, stripe/egr-1 (Takayanagi-Kiya et al., 2023). That study used a GAL4 driver for the stripe/egr-1 gene to label and functionally manipulate activated neurons. In contrast, our approach is based purely on detecting expression of the IEG mRNA using a state-of-the-art amplification method, HCR v3.0 (Figure 1-figure supplement B) (Choi et al., 2018). The HI-FISH method enables us to test essentially any IEG of interest, rather than only those IEGs for which GAL4 drivers are available. Furthermore, by using direct mRNA detection rather than an IEG-GAL4 driver, a series of GAL4 or split-GAL4 lines can be combined with our approach to characterize the identity of IEG-labeled neurons. Thus, the two methods complement each other.

Materials and methods

Fly stocks

A list of the full genotypes for the flies used in this research is in Supplementary file 1. R15A01-iLexA (improved LexA) was generated for this study. Briefly, the R15A01 enhancer fragment (Jenett et al., 2012) was cloned into pBPnlsLexA::p65::GADUw (Chiu et al., 2021).

20xUAS-IVS-Syn21-Chrimson::tdTomato3.1 (attP2) and 13xLexAop2-IVS-Syn21-Chrimson::tdTomato3.1 (su(Hw)attP5) were obtained from G. Rubin (Janelia Research Campus). 0273-GAL4 (Burke et al., 2012) was obtained from S. Waddell (The University of Oxford). fruGAL4 (Stockinger et al., 2005) was from B. Dickson (Janelia Research Campus).

The following strains were obtained from Bloomington Drosophila Stock Center (Indiana University); R15A01-AD (attP40), R71G01-DBD (attP2), R41A01-GAL4 (attP2), OK107, Ddc-GAL4, 20xUAS-IVS-jGCaMP7b (VK00005).

Optogenetic activation in behaving flies

A detailed description of the setup used for optogenetic activation was described previously (Inagaki et al., 2014; Watanabe et al., 2017). Flies were raised at 25℃ and collected on the day of eclosion and kept in the dark for 7-10 days on the fly media containing 0.4 mM all trans-Retinal (Merck, St. Louis, MO). The experiments were performed with the “8-well” chamber (16 mm diameter x 10 mm height) with IR backlight. (855 nm, SOBL-6×4-850, SmartVision Lights, Norton Shores, MI). A 655 nm 10 mm Square LED (Luxeon Star LEDs Quadica Developments, Brantford, Canada) was used for photostimulation. The stimulation protocols for each experiment are described in the figures or figure legends.

Behavioral Assays

Flies were maintained at 9AM:9PM Light:Dark cycle. For male-male aggression assays and male-female courtship assays, the “8-well” acrylic chamber was used as described previously (Inagaki et al., 2014; Watanabe et al., 2017). The walls of the chamber were coated with a fluoropolymer resin (PTFE-30, Insect-a-Slip) (BioQuip Products, Rancho Dominguez, CA). The clear top ceiling plate was coated with Sigmacote (Merck, St. Louis, MO) to prevent flies from walking on the surfaces. The floor was covered with a layer of apple juice/sucrose-agarose food. The arena was illuminated with IR backlighting.

Two-photon calcium imaging

Calcium imaging with optogenetic activation was performed as described (Inagaki et al., 2014; Jung et al., 2020). A deep red (660 nm) fiber-coupled LED (Thorlabs, Newton, NJ) with a band-pass filter (660 nm, Edmund Optics, Barrington, NJ) was used for light source to activate Chrimson.

Flies were collected on the day of eclosion and kept in the dark on the fly media containing 0.4 mM all trans-Retinal (Merck, St. Louis, MO). 7-10 days old flies were cold-anesthetized using a temperature-controlled stage and mounted on a plastic plate with a small window using wax. The cuticle of the dorsal side of the head was removed by forceps. Flies were then placed beneath the objective.

Fly saline (108 mM NaCl, 5 mM KCl, 4 mM NaHCO3, 1 mM NaH2PO4, 5 mM trehalose, 10 mM sucrose, 5 mM HEPES, 0.5 mM CaCl2, 2 mM MgCl2, pH = 7.5) was used to bathe the brain during imaging.

Immunohistochemistry and HCR fluorescence in situ hybridization (FISH)

In situ HCR was performed as previously described with minor modifications (Choi et al., 2018, also see https://www.moleculartechnologies.org/supp/HCRv3_protocol_generic_solution.pdf). For the hybridization buffer, formamide was substituted with an equal volume of 8M urea solution for better preservation of tissue morphology (Sinigaglia et al., 2018).

To process multiple individuals, we collected flies (∼ 8 individuals at once), soaked them in ice-cold ethanol to remove the wax on the cuticle, and opened the head cuticle of each fly with minimal time-lag among samples in the ice-cold fixative (4% paraformaldehyde in PBS). After brief fixation, the brain from each individual was dissected and fixed for 2 hours at 4℃.

For combination of immunohistochemistry (IHC) and HCR FISH, we performed IHC as described previously before HCR FISH procedure in RNase free condition (Watanabe et al., 2017). For antibody incubation, we added RNasin Plus (5% of final volume, Promega, Madison, WI). After immunostaining, the samples were post-fixed with 4% formaldehyde in PBS and proceeded for HCR. HCR probes were purchased from Molecular Technologies (https://www.moleculartechnologies.org/). The samples were mounted in 88% (w/v) Histodenz in 0.02 M Phosphate buffer (pH 7.5 with NaOH) with 0.1% Tween-20 and 0.01% sodium azide (Yang et al., 2014).

The antibodies used were as follows: rat anti-DN Cadherin (1:50, DN-EX #8, developed by T. Uemura, obtained from Developmental Studies Hybridoma Bank, University of Iowa, Iowa City, IA), mouse anti-Bruchpilot (1:50, nc82, developed by E. Buchner, obtained from Developmental Studies Hybridoma Bank, University of Iowa, Iowa City, IA), rabbit anti-GFP (1:1,000, A11122, Thermo Fisher Scientific, Waltham, MA), Goat anti-Mouse IgG (H+L) Alexa Fluor 488 (1:1,000, A-11001, Thermo Fisher Scientific, Waltham, MA), Goat anti-Mouse IgG (H+L) Alexa Fluor 568 (1:1,000, A-11004, Thermo Fisher Scientific, Waltham, MA), Goat anti-Mouse IgG (H+L) Alexa Fluor 633 (1:1,000, A-21050, Thermo Fisher Scientific, Waltham, MA), Goat anti-Rabbit IgG (H+L) Alexa Fluor 488 (1:1,000, A-11008, Thermo Fisher Scientific, Waltham, MA), Goat anti-Rabbit IgG (H+L) Alexa Fluor 488 (1:1,000, A-11011, Thermo Fisher Scientific, Waltham, MA), Goat anti-Rabbit IgG (H+L) Alexa Fluor 633 (1:1,000, A-21070, Thermo Fisher Scientific, Waltham, MA)

Confocal serial optical sections were obtained with a Fluoview FV3000 Confocal Microscope (Olympus, Tokyo, Japan).

For the Hr38 exonic probe, we first measured average fluorescence intensity within the cell boundary using Fiji (Schindelin et al., 2012; Schneider et al., 2012) and calculated the relative intensity to the background signals and then defined the positive cells whose signals were above 3σ.

For image registration, brain images were registered to T1 template brain (Yu et al., 2010) using the CMTK registration plugin of Fiji (https://flybrain.mrc-lmb.cam.ac.uk/dokuwiki/doku.php?id=warping_manual:registration_gui). To create average images for each condition, MATLAB was used to generate averaged images of each z-plane from different individuals. For voxel-based heat map analysis of the Hr38 expression, the median filter (pixel size = 10) in Fiji was applied to the averaged images. Fiji and Fluorender software (Yong Wan et al., 2009) were used to create z stack images.

catFISH

To prepare target flies, small iron filings were attached on the dorsal side of the thorax of 6-8 days old male or female flies using UV-activated glue. This allowed us to remove the target fly from a behavioral arena through a small hole using a neodymium magnet.

6-8 days single-housed male flies were used for the experimental flies. A male fly was introduced in the behavioral arena and allowed to habituated to the new environment for ∼ 3 hours. The first target fly was introduced into the arena, and the experimental fly was exposed to the target for 20 mins. After the exposure, the target fly was removed from the arena as describe above. The experimental fly was given a 30-min interval before the second target was introduced. After 20 min exposure, the second target was removed, and Hr38 expression was analyzed 10 min after the target fly removal (Figure 4A and B). Neurons with a clear dot signal were counted as positive for the Hr38 intronic probe.

Statistical Analysis

Statistical analyses were performed using MATLAB (MathWorks, Natick, MA) and Python (Python Software Foundation). The detailed information including the number of samples, the statistical method, and the p value for each experiment is indicated in the figure legends.

Calcium imaging data analysis

ROIs (region of interest) corresponding to nerve bundle or cell bodies of labeled neurons were manually selected and the fluorescence intensity in the ROIs was measured using Fiji. The average fluorescence intensity during the first 15 s of image acquisition was used as the baseline to calculate the ΔF/F. The mean of ΔF/F during the indicated periods was calculated and visualized using Python. For cell body analysis, we defined the positive response if the average ΔF/F during the photostimulation was statistically higher than the baseline (p < 0.01). The strong response neurons (ΔF/F > 0.4) and the intermediate neurons (0.15 < ΔF/F ≤ 0.4) were defined based on the average ΔF/F value during the stimulation, respectively.

Author Contributions

Kiichi Watanabe, Conceptualization, Data curation, Formal analysis, Investigation, Methodology, Resource, Software, Visualization, Writing - original draft, Writing - review and editing; Hui Chiu, Resource; David J. Anderson, Conceptualization, Funding acquisition, Methodology, Project administration, Supervision, Writing - original draft, Writing - review and editing.

Supporting information

eLife Assessment

The work addresses an important methodological aspect by optimizing an activity-dependent labelling of neural circuits in behaving flies. The authors provide convincing evidence to support the broad applicability of this method. However, a more comprehensive description of the methodology would greatly enhance its dissemination and adoption. Additionally, the authors successfully implement the method, providing solid evidence for the activity-dependent labelling of P1 neurons during aggression and courtship.

Reviewer #1 (Public Review):

Summary:

The authors have nicely demonstrated the efficiency of the HCR v.3.0 using hr38 mRNA expression as a marker of neuronal activity. This is very important in the Drosophila neuroscience field as in situ hybridization in adult Drosophila brains has been so far very challenging to do and replicate. However, this method has been described before [Choi et al., (2018)] and, if I understand correctly, is now the property of the non-profit organization molecular Technologies, who are the ones responsible for designing the probes (the sequences are not provided). Here the authors present their work as a description of a new method, called HI-FISH. However, I do not consider this as a new method but rather an application, a "proof of principle" that HCR v.3.0 can be done even on challenging tissues such as the adult Drosophila brain. Hence, if HCR v3.0 is sensitive enough and powerful enough to be used as a marker of neuronal activity, we can use it, for other neurobiological purposes, using other gene probes.

To demonstrate the efficiency of HI-FISH, the authors have addressed two biological questions. The first one addressed whether specific groups of neurons, known to trigger specific behaviours (here courtship and/or aggression) are indeed activated by the behavioural context they can promote. In other words: is the behavioural output of these neurons also a trigger for their activation? The second question addressed whether this method is powerful enough to distinguish two subgroups of a class of neurons called P1 known to be involved in the two behaviours considered. In other words, are the same P1 neurons that promote aggression and courtship?

Strengths: The demonstration of the efficiency of the method is very convincing and well-performed. It gives the will for the reader to apply the method to their own subject.

Weakness: The pictures provided for HI-FISH and catFISH do not corroborate with the quantification and therefore I am not convinced about the author's biological interpretation of their data. See below for details.

Previously, using a split-gal4 line to restrict the Gal4 expression to a subset of P1 neurons, the authors have shown that these particular neurons when activated can trigger both aggressivity and courtship behaviour [Hoopfer et al., 2015]. The P1 neurons are composed of about 20 FruM neurons/hemibrain but are part of a bigger group that comprises about the same number of Fru- neurons that seem to be exclusively aggression-promoting neurons [Koganezawa et al., 2016]. Hence, this group of 40 neurons (pC1 neurons) contains aggressive-promoting neurons, courtship-promoting neurons, and perhaps neurons that can do both. Therefore, to address the first question, the authors compared hr38 expression in different groups of neurons, with a focus on subgroups, under different social contexts. While there is no ambiguity concerning the function of the Tk neurons as being exclusively aggressive-promoting neurons [Asahina et al., 2014], it is less clear when we look at the pC1 neurons. This is particularly evident for the P1a neurons which have been shown to be ambiguous as they can promote both aggression and courtship. For example, while optogenetic activation of these neurons promotes hr38 expression (Fig. 3D and fig sup. 4), it is less clear in the pictures to determine whether these specific P1a neurons labeled by the split-gal4 line are specifically activated by an aggressive behavioural context or a courtship behavioural context (Fig1, supp. 2 and fig. 4). Furthermore, the pictures chosen do not reflect the reality of the quantification (Fig. 2 B-D compared to sup. 2 or fig. 4C compared to fig. 4D) and therefore the authors conclusion. Because the drivers used are only expressed by a small subset of a larger population, I believe it would be more informative to separate in the quantification between the Gal4-expressing neurons and the non-expressing ones. Notably, based on the pictures provided, it looks like more P1 neurons (or rather pC1 neurons) are activated by a male-male encounter than by a male-female encounter. On the other hand, the splitGal4+ P1a seem to be more responsive to a courtship behavioural context (2/6 P1a neurons express hr38 in a courtship behavioural context while 0/9 _if we mentally abstract the increase of the background signal compared to the control picture_ seem to express hr38 in an aggression behavioural context). Hence, while activation of this P1asplit-Gal4 can promote both aggressive behaviour and courtship behaviour [Hoopfer et al., 2015], the authors didn't provide clear evidence (pictures not corroborating the quantification) that these specific small subpopulation of neurons are activated by one or the other or both behavioural conditions. Therefore my suggestion of differentiating in the quantification between the Gal4+ neurons from the others in the same local area.

Fig. 3, suppl. 3: In this section the authors addressed the question of whether the HI-FISH can be used to identify the downstream targets of this subpopulation. As positive controls of known downstream targets, the authors looked at the pCd population which they recently published as being an indirect downstream target of the P1a neurons [Jung et al., Neuron 2020]. They identified the Kenyon cells and a group of dopaminergic neurons, the PAM neurons as being activated by the P1a neurons. To confirm the increase of hr38 expression is indeed the result of a neuronal response of these neurons to the P1a activation, the authors used a different strategy used by them and others before. Using Gcamp signal to monitor the neuronal response of the presumably downstream targets the authors activated the P1a neurons using optogenetic (chrimson). It is important to note that they have previously shown that depending on the frequency of the light pulses activation of the P1a neurons can trigger only aggression, both aggression and wing extension or only wing extension [Hoopfer et al., eLife 2015]. Here the authors use 50Hz which is a frequency that leads to wing extension during the stimulation and aggressive behaviour at the offset of the stimulus [Hoopfer et al., eLife 2015]. Interestingly, the Gcamp experiment shows activation of the Kenyon cells and the PAM neurons but this activity is not maintained when the stimulus is turned off, suggesting that these neurons are activated during a courtship context rather than an aggressive behavioural context. I think it would be nice to see in which behavioural context the Kenyon cells and PAM neurons are activated (hr38 expression in the different behavioural context using the corresponding Gal4).

Fig.4 and supp.4: The demonstration that the catFISH can now be done in Drosophila brain with a new in situ method was nicely performed. Notably, the intronic Hr38 probe appears to be an excellent marker for recent neuronal activation. However, while the optogenetic activation of the P1a neurons used to quantify the time lapse for both probes nicely distinguishes between nuclear and cytoplasmic exonic hr38, it is very difficult to use the localization of this probe in the experimental setup the authors used. Also, With their setup, I would simply use the frequency of intronic hr38 as a marker of recent activation correlating or not with the frequency of exonic hr38 marker (present in both conditions first and second encounter). This is important as this experiment addresses the second biological question. Once again, the pictures chosen absolutely do not corroborate the quantification. For example, the picture of the double encounter with the same gender male-male context clearly shows a higher number of cells that are hr38INT positive (and therefore nuclear) than the picture of the female-female context (Fig. 4C), and thus even if we only considered the P1asplit-Gal4 positive cells. In the male-male picture, 5/6 P1a cells have the Hr38INT marker while the presence of this marker is debatable in the female-female context. Especially, in some of the cells these magenta dots appear to be localized in the cytoplasm, suggesting a non-specific signal. Therefore, I would suggest to quantify the percentage of Hr38INT positive cells as the only marker for recent activation and the relative level of Hr38EXN immunostaining, and this among the P1asplit-Gal4 positive cells and the gal4- ones. A high Hr38EXN level associated with the presence of hr38INT would indicate that the cell has been activated during both encounters, while a lower hr38EXN with no hr38INT would suggest only an activation during the 1st behavioural context. Finally, a lower hr38EXN associated with the presence of hr38INT would suggest the opposite, an activation only during the 2nd behaviour.

Overall, by only looking at the pictures provided, I would conclude that the HCR applied with the hr38 probes seems to efficiently work and is usable to address the question of whether a specific group of neurons are indeed activated by a specific social behavioural context. However, I would also conclude that this technique nicely demonstrated that flies are not robots and that even in a "simple" organism model such as Drosophila melanogaster individual variability is present among a group of neurons. Hence, only the quantification of the gal4-expressing neurons in comparison with their neighbor neurons known to belong to the same functional group, would allow a conclusion toward a specificity of contextual response. Therefore, although activation of a small group of neurons can be enough to trigger a specific behaviour that shouldn't happen under a certain environmental context [Hoopfer et al., eLife 2015], the results presented here suggest that we should, using this method, considering the response of the neighbour cells of the Gal4+ ones. Although currently, the quantification of the author's data does not allow such analysis, to strengthen the author's argumentation, I would distinguish in their quantification between gal4+ from the others (Fig. 2 and 4). Furthermore, I am not certain that the distinction between cytoplasmic and nuclear hr38EXN is 100% feasible (based on the pictures provided). I would instead for the hr38EXN marker only use the relative intensity (Fig. 4D).

Reviewer #2 (Public Review):

Summary:

Watanabe et al establish a novel method for the activity-dependent labeling of neural circuits in flies. While activity mapping of neurons that are active during specific behaviors is widespread in rodents, the application of this method to fly circuit neuroscience is limited, mainly due to technological challenges. Thus, the present study addresses a timely problem. To do so, they apply the in situ hybridization amplification method called Hybridization Chain Reaction v. 3.0 (Choi et al. 2018) to the adult fly brain in order to visualize the expression changes of the immediate early gene (IEG) Hr38 under different types of social contexts. The conclusions of this paper are mostly very well supported by data but it would strongly benefit from additional methodological details as well as additional controls, in particular for the HI-catFISH experiments.

Strengths:

The major strength of this method is its versatility and sensitivity. It can be applied to a wide variety of biological questions and assess the dynamic transcriptional regulation of an unlimited number of genes with a high signal-to-noise ratio. It will be therefore useful to many research labs working on different biological questions.

Weaknesses:

Although the paper has great strengths in principle, the major weakness is the calibration of the temporal resolution of HI-CatFISH in Figure 4 and Figure Supplement 4. According to Figure Supplement 4C, close to 100% of the Hr38-positive cells are already labeled with the exonic probe 30min post-stimulation, which is not reflected in Figure 4B (there, the expression level of the exonic probe peaks 60min post-induction) and may have profound implications for the interpretation of the results. The present manuscript would strongly benefit from additional controls, such as the quantification of the intronic and exonic Hr38 probes after either only the 1st or 2nd social context but at the same timepoint than if two consecutive social contexts were tested.

Author response:

Reviewer #1:

Response to Public Review

We thank the reviewer for taking the time to carefully read our paper and to provide helpful comments and suggestions, most of which we have incorporated in our revised manuscript. One of this reviewer’s (and reviewer #2’s) main concerns was that the confocal images provided in some cases did not appear to reflect the quantitative data in the bar graphs. These images were provided only for illustrative purposes, to give the reader a sense of what the primary data look like. The reviewer may not have appreciated that the quantitative data reflect counts of RNA smFISH signals (dots) in hundreds of cells collected through z-stacks comprising multiple optical sections in multiple flies for each condition For example, in P1a control condition (in Figure 2A), we have analyzed 135 neurons from 8 individuals. There, the number of z-planes ranged from 3 to 8 per hemisphere. It is generally not possible to find a single confocal section that encompasses quantitatively the statistics that are presented in the graphs. Presenting the data as an MIP (Maximum Intensity Projection, i.e., collapsed z-stack) in a single panel would generate an image that is too cluttered to see any detail. We have now included, for the reader’s benefit, additional example confocal sections in both a z-stack and from the opposite hemisphere, in Supplemental Figure S4D. We have also inserted clarifying statements in the text on p. 7 (lines 154-156).

Another suggestion from Reviewer #1 is that "it would be more informative to separate in the quantification between the GAL4-expressing neurons and the non-expressing ones" based on the presented pictures where more non-P1a neurons (that the reviewer speculates may be pC1-type neurons) are activated by a male-male encounter than by a male-female encounter, while the P1a-positive neurons seem to be more responsive during courtship behavior. In this paper, we were not looking at pC1 neurons and did not try to answer which neuronal population(s) outside of the P1a population is/are responsible for aggression and/or courtship. Rather, we focused on P1a neurons and addressed whether P1a neurons that induce both aggression and courtship behavior when they are artificially activated (Hoopfer et al. 2015) are also naturally activated during spontaneous performance of these two social behaviors. However, this result did not exclude the possibility that P1a neurons were inactive during naturalistic courtship or aggression. Our data in the current manuscript provide further experimental evidence in support of the idea that P1a neurons as a population play a role in both of these behaviors. Moreover, we provided data identifying P1a neurons activated only during aggression or during courtship (or both). However this does not exclude that pC1 or other neighboring populations are activated during aggression as well (See also the response to 'Recommendations For The Authors' and text lines 151-154).

In Figure 3, we used opto-HI-FISH to identify candidate downstream targets (direct or indirect) of P1a neurons. We used 50 Hz Chrimson stimulation to activate P1a neurons to induce expression of Hr38 and identified Kenyon cells in the mushroom body (MB) and PAM neurons (as well as pCd neurons) as potential downstream targets of P1a cells. In Figure 3 – supplement we performed calcium imaging of KCs and PAM neurons in response to P1a optogenetic stimulation to confirm independently our results from the Hr38 labeling experiments. That control was the purpose of that supplemental experiment.

Based on those imaging data, the reviewer asked the further question of which [natural] behavioral context induces Hr38 expression in these populations (i.e., mating or aggression). This question is reasonable because our calcium imaging data (Figure 3-supplement) showed that both Kenyon cells and PAM neurons are active only during photo-stimulation of P1a neurons. Our previous behavioral studies (Inagaki et al., 2014; Hoopfer et al., 2015) showed that 50 Hz photo-stimulation of P1a neurons in freely moving flies induced unilateral wing extension during stimulation, while aggression was observed only after the offset of the stimulation (Hoopfer et.al., 2015). Based on the comparison of those behavioral data to the imaging results in this paper, the reviewer suggested that Kenyon cells and PAM neurons are activated during courtship rather than during aggression. This is certainly a possible interpretation. However it is difficult to extrapolate from behavioral experiments in freely moving animals to calcium imaging results in head-fixed flies, particularly with response to neural dynamics. Furthermore, Hr38 expression, like that of other IEGs (e.g., c-fos), may reflect persistently activated 2nd messenger pathways (e.g., cAMP, IP3) in Kenyon cells and PAM neurons that are not detected by calcium imaging, but that nevertheless play a role in mediating its behavioral effects. We still do not understand the mechanisms of how optogenetic stimulation of P1a neurons in freely behaving flies induces aggression vs. courtship behavior. Although 50 Hz stimulation of P1a neurons does not induce aggressive behavior during photo-stimulation, it is possible that this manipulation activates both aggression and courtship circuits, but that the courtship circuit might inhibit aggressive behavior at a site downstream of the MB (e.g., in the VNC). Once stimulation is terminated and courtship stops the fly would show aggressive behavior, due to release of that downstream inhibition (see Models in Anderson (2016) Fig 2d, e). In that case, there would be no apparent inconsistency between the imaging data and behavioral data. We agree that the reviewer's question is interesting and important but we feel that answering this question with decisive experiments is beyond the scope of this manuscript.

Finally, Reviewer #1 suggested a method to evaluate the Hr38 signals in the catFISH experiment of Figure 4. We appreciate their suggestions, but the way that we evaluated the Hr38 signals was basically the same as the way the reviewer suggested. We apologize for the confusion caused by the lack of detailed descriptions in the original manuscript. We have now revised the methods section to explain more clearly how we define the cells as positive based on Hr38EXN and Hr38INT signals.

Response to Recommendations for the authors:

“To strengthen the author's argumentation, I would distinguish in their quantification between gal4+ from the other [classes of neighboring neurons]” (Fig. 2 and 4).”

Our focus in this paper was to ask simply whether P1a neurons are active or not active during natural occurrences of the social behaviors they can evoke when artificially activated. We did not claim that they are the only cells in the region that control the behaviors. It is not possible to compare their activation to that of 'other' cells neighboring P1a neurons without a separate marker to identify those cells driven by a different reporter system (e.g., LexA). This in turn would require repeating all of the experiments in Figs 2 and 4 from scratch with new genotypes permitting dual-labeling of the two populations by different XFPs, and quantifying the data using 4-color labeling. We respectfully submit that such curiosity-driven experiments, while in principle interesting, are beyond the scope of the present manuscript. However, we have inserted text to acknowledge the possibility that the aggression-activated Hr38 signals in P1a- cells neighboring P1a+ cells may correspond to other classes of P1 neurons (of which there are 70 in total) or to pC1 cells. Changes: Text lines 151-154.

“if the magenta dot is outside of the nuclei I would not count this as positive also the size of the dot seems to be a good marker of the reality of the signal). I would measure the intensity of the hr38EXN. A high Hr38EXN level associated with the presence of hr38INT would indicate that the cell has been activated during both encounters, while a lower hr38EXN with no hr38INT would suggest only an activation during the 1st behavioural context. Finally, a lower hr38EXN associated with the presence of hr38INT would suggest the opposite, an activation only during the 2nd behaviour.”

We agree that there are some tiny dot signals with hr38 INT probe that are more likely the background signals. We only counted the INT probe signals as positive when the cells had a clearly visible dot and also co-localize with the exonic probe's signal, as primary (un-spliced) Hr38 transcripts in the nucleus should be positive for both EXN and INT probes. Regarding the reviewer’s latter comments, we agree with their interpretation of the catFISH results and that is how we interpreted them originally. We measured the intensity of hr38EXN expression and defined hr38EXN-labeled cells as “positive” when the relative intensity was 3σ >average, a stringent criterion. In the revised manuscript, we added more detailed information in the methods section regarding our criteria for defining cell types as positive.

“Knowing that the P1a neurons (using the split-gal4) can trigger only wing extension when activated by optogenetic 50Hz, I would test to which behavioral context the MB neurons and the PAM neurons positively respond to.”

As we answered in 'Response to Public Review,' our opto-HI-FISH experiments identified Kenyon cells in the mushroom body (MB) and PAM neurons (as well as pCd neurons) as potential downstream targets of P1a cells, using Hr38 labeling. The purpose of the calcium imaging experiment in Figure 3 – supplement was to confirm the P1a-dependent activation of KCs and PAM neurons using an independent method. In that respect this control experiment was successful in that methodological confirmation. The reviser raised an interesting question about how our calcium imaging experiments relate to our behavioral experiments, in terms of the dynamics of KC and PAM activation. A recent publication (Shen et al., 2023) revealed that courtship behavior has a positive valence and that activation of P1 neurons mimics a courtship-reward state via activation of PAM dopaminergic neurons. Therefore, it is reasonable to think that PAM neurons (and Kenyon cells as downstream of PAM neurons) are activated during female exposure. However those data do not exclude the possibility that inter-male aggression is also rewarding in Drosophila males, as it has shown to be in mice. This is an interesting curiosity-driven question that has yet to be resolved. Therefore, as mentioned in the 'Response to Public Review,' we feel that the additional experiment the reviewer suggests is beyond the scope of our manuscript.

Changes: None.

Minor comments:

“Please provide different pictures from main fig2 and sup2 for the three common conditions (control, aggression, and courtship).”

The data set for Figure 2 and Figure 2 supplement are from the same experiment. Because of the limited space, we just presented the selected key conditions ('Control', 'Aggression', and 'Courtship') in the main figure and put the complete data set (including these three key conditions) in the supplemental figure.

Changes: None

“Please, provide scale bars for the images.”

Also, Reviewer #2 commented, 'Scale bars are missing on all the images throughout the main and supplementary figures.'

We have now added scale bars for each figure.

“Fig.1: “Is the chrimsonTdtom images from endogenous fluorescence? It is not said in the legend and anti-dsred is not provided in the material and method while anti-GFP is.”

We are sorry for the confusion and thank the reviewer for raising that question. The signals were native fluorescence, and we have now added that information to the figure legend.

P7: "As an initial proof-of-concept application of HI-FISH, we asked whether neuronal subsets initially identified in functional screens for aggression-promoting neurons (Asahina et al., 2014; Hoopfer et al., 2015; Watanabe et al., 2017) were actually active during natural aggressive behavior. These included P1a, Tachykinin-FruM+ (TkFruM), and aSP2 neurons". Please put the references to the corresponding group of neurons listed. For example: "These included P1a neurons [Hoopfer et al., 2015]".

We have now added these references.

P9: "Optogenetic and thermogenetic stimulation experiments have shown that that P1a interneurons can promote both male-directed aggression and male- or female-directed courtship" typo

We appreciate the reviewer for catching this error and have corrected the text.

(P10:" To validate this approach, we first asked whether we could detect Hr38 induction in pCd neurons, which were previously shown by calcium imaging to be (indirect) targets of P1a neurons". Reference [Jung et al., 2020]

We have now added this reference.

Fig. 4A: Put the time scale on the diagram (3h adaptation-20min-30min rest-20min-10min rest-collect)

We have now added the time scale in Figure 4A.

Reviewer #2:

Response to Public Review:

We thank the reviewer for their helpful comments and suggestions. We have addressed most of them in our revised manuscript. The main concern of Reviewer #2 was the temporal resolution of the HI-catFISH experiment shown in Figure 4 and Figure 4-Supplement. Our original manuscript illustrated temporal patterns of Hr38EXN and Hr38ITN signals concomitant with different behavioral paradigms (Figure 4B). The reviewer pointed out that the illustrated experimental design does not reflect the actual data shown in Figure 4-Supplement A-C. We believe this issue was raised because we drew the temporal pattern of Hr38EXN signals in Figure 4B based on the intensity of Hr38EXN signals (Figure 4-Supplement B) rather than based on the % number of positive cells (Figure 4-Supplement C). We have now revised the schematic time course of Hr38EXN signals in Figure 4B using the % of positive cells. We believe this change will be helpful for readers to understand better the experimental design since we used the % of positive cells to identify patterns of P1a neuron activation during male-male vs. male-female social interactions in Figure 4D. Another suggestion from Reviewer #2 was to add additional controls, such as the quantification of the intronic and exonic Hr38 probes after either only the first or second social context exposure. In response, we have now added the data from only the first social context (Figure 4C, and 4D, right column). These new data provides evidence that there are essentially no detectable Hr38INT signals 60 minutes later without a second behavioral context, while Hr38EXN signals are still present at the time of the analysis. Unfortunately, we are not able to provide the converse dataset with the second behavioral context only to show that Hr38 INT signals are detected. On this point, we call the reviewer’s attention to Figure 4-supplement-S4A-C, which show that the INT probe signals are detectable at 15 and 30 minutes following stimulation, but not at 60 minutes. In the experiment of Fig. 4B, flies are fixed and labeled for Hr38 30 minutes after the beginning of the second behavior, conditions under which we should obtain robust INT signals (as observed). EXN signals are also expected at 30 minutes because the primary (non-spliced) RNA transcript detected by the INT probe also contains exonic sequences.

Response to Recommendations for the authors:

Given that the development of in situ HCR for the adult fly brain is so central to the present manuscript, I think that the methods section describing the HCR protocol can be significantly improved. In particular, the authors should fully describe the in situ HCR protocol including the 'minor modifications' they refer to, and define how they calculate the 'relative intensity to the background'.

We appreciate the reviewer’s suggestion. We have now revised the methods section to describe the procedure in more detail. Also, we will submit a separate document describing the HI-FISH protocol.

Note: The authors refer to a recently published paper by Takayanagi-Kiya et al (2023) describing activity-based neuronal labeling using a different immediate early gene, stripe/egr-1. The authors state the following: 'That study used a GAL4 driver for the stripe/egr-1 gene to label and functionally manipulate activated neurons. In contrast, our approach is based purely on detecting expression of the IEG mRNA using..'. Takayanagi-Kiya et al. (2023) also use in situ mRNA detection of the IEG stripe/egr-1 and not only a GAL4 driver system. This claim should be modified and the paper should be cited in the introduction of the present paper.

We have now cited the paper in the Introduction and have modified and moved the description originally in 'Note' section to Discussion (text lines: 392-404) as the reviewer requested. We have emphasized the difference between the two approaches for comparing neuronal activities during two different behaviors within the same animal. Takayanagi-Kiya used GAL4/UAS and stripe protein expression with immunohistochemistry to analyze neuronal activities during two different behaviors, while we exclusively analyzed Hr38 mRNA expression for this purpose, using intronic and exonic Hr38 probes. This approach made it possible to perform catFISH with higher temporal resolution and also allows extension of our approach to other IEGs for which antibodies are not available.

Please specify the nature of the iron fillings in the methods section.

We added a detailed description in the methods section, including the catalog number.

In Figure 1B, the authors may add a dashed outline to the regions magnified in 1C so that readers can more easily follow the figures. Moreover, it would be informative to see a more detailed quantification of the number of Hr38-positive cells in different brain regions marked by Fru-GAL4.

We have now added the whole brain images for each condition in Figure 1C and also quantitative data in Figure 1-Supplement C, as the reviewer suggested.

In the middle right aggression panel of Figure 2A, it looks as if one P1a neuron is not outlined.

We have carefully examined other z-planes through this region and based on those data have concluded that the signals mentioned by the reviewer are neurites from neurons labeled in other z-planes.

Changes: None.

The images in Figure 2A can be again found in Figure Supplement 2A, yet the number of neurons analyzed suggests the quantification was performed from different samples. The images in Figure Supplement 2A should be either changed or it should be explained as to why the images are the same yet the numbers in the legend are different.

We apologize for the confusion. Figure 2 and Figure 2-Supplement are from the same experiment. To avoid clutter we illustrated three key conditions ('Control,' 'Aggression,' and 'Courtship') in the main figure. The reason why the numbers in the legend are different is that the purpose of presenting Figure 2-Supplement B-D was to determine whether there were differences in the intensity of Hr38 FISH signals in the neurons considered as 'positive' in different conditions. Therefore, the numbers described in Figure 2-Supplement legend are derived only from those neurons that were considered Hr38-positive, while the numbers in Figure 2 include all neurons analyzed. We have now added notes to explain this in the Figure 2 – supplement legend.

The panels of the quantification of the Hr38 relative intensity in Figure 2B/C/D are very difficult to read, ideally, they should be plotted as in Figure Supplement 2B/C/D.

The graphs in Figure 2B-D (upper) show data from all GFP-labeled cells scored, including cells defined as 'negative' or 'borderline.' In contrast, the graphs in Figure 2-supplement show the relative Hr38 signal intensity in those GFP neurons defined as positive based on the analysis in Fig. 2B. If we were to plot the data in Fig. 2B (upper) as box plots (like that in Figure-2-supplement), we would see either a skewed (only negative cells) or a bimodal distribution (one around the negative population and the other around the positive population); the shapes of these distributions would likely be hidden in the box-whisker plots format. Therefore, we prefer to plot all of the data points as we did in the original manuscript. However, we agree that the data points in the original manuscript were hard to read. We therefore changed the format of the datapoints from blurry dots to open circles with clear solid lines.

In Figure 2B/C/D, please specify in the figure legend what 'grouped in categories according to character' means.

We used letters to mark statistically significant differences (or lack thereof) between conditions. Bars sharing at least one common letter are not significantly different. If they do not share any letter, they are significantly different. For example, Aggression: bc vs. Dead: bc, means no difference. Aggression: bc vs. No Food: b, or Aggression: bc vs. Courtship: c also means no difference between Aggression and each of the two other conditions. However, 'No Food: b' and 'Courtship: c' have no common letter, meaning they are different. This is a standard method for showing statistically comparisons among multiple bars without lots of asterisks and horizontal bars cluttering the figure, and we have revised the legend to clarify what each letter means. We have also removed the color shading in Figure 2 B-D as it may have been confusing.

A quantification of the number of Hr38-positive neurons and Hr38 relative intensity during the entire time course would be informative in Figure 3D.

Although the data set for this figure is different from that for Figure 4-Supplement A-C, the main claim is the same. Therefore, Figure 4 - Supplement essentially provides the information that the reviewer suggested. However, we also reanalyzed the data set used for the original Figure 3D and evaluated % positive cells at the 30-minute time point and have now added that number in the figure legend.

In the legend of Figure 3D, it says '..The expression level reaches its peak at 30-60min', yet I don't see timepoints beyond 60min. Please rephrase or add additional timepoints.

We apologize for the error. We have rephrased the text.

Figure Supplement 3A/D: please add an outline or a schematic figure to better understand where the imaging is performed.

We added illustrated schemas next to the title of each experiment (P1->PAM neurons (bundle) and P1 -> Kenyon cells (bundle)).

Figure Supplement 3C/F: please add information about the statistical test to the corresponding figure legend.

We have added a phrase to describe the test used.

Figure Supplement 3G/H/I/J: motion artifacts can potentially strongly affect the performed analysis given that cell bodies are very small and highly subjected to motion. Can the authors comment on how they corrected for motion?

We have now described how we corrected for motion artifacts in the Methods section.

Figure 4C/D: It seems as if the representative images don't reflect the quantification, e.g., in the male -> female panel, close to 100% of the neurons are positive for the exonic probe as opposed to approx. 40% in the bar graph.

Please see our response to this issue in the 'Response to Public Review (Reviewer #1)'.

Additional controls should be included in Figure 4C in order to assess the temporal resolution of HI-CatFISH more in detail (see 'Weaknesses').

We have also answered this in the 'Response to Public Review'.

The authors should adjust the scheme in the main Figure 4B to reflect the data presented in Figure S4A and C. For instance, the peak for the intronic version is observed at 15 minutes, while at 30 minutes, both the exonic and intronic signals show an equal level of signal.

We have addressed this issue in the 'Response to Public Review'.

We thank the reviewers again for their helpful comments and hope that with these changes, the manuscript will now be acceptable for official publication in eLife.

Version 2

Abstract

Abstract

Monitoring neuronal activity at single-cell resolution in freely moving Drosophila engaged in social behaviors is challenging because of their small size and lack of transparency. Extant methods, such as Flyception, are highly invasive. Whole-brain calcium imaging in head-fixed, walking flies is feasible but the animals cannot perform the consummatory phases of social behaviors like aggression or mating under these conditions. This has left open the fundamental question of whether neurons identified as functionally important for such behaviors using loss- or gain-of-function screens are actually active during the natural performance of such behaviors, and if so during which phase(s). Here we perform brain-wide mapping of active cells expressing the Immediate Early Gene hr38 using a high-sensitivity/low background FISH amplification method called HCR-3.0. Using double-labeling for hr38 mRNA and for GFP, we describe the activity of several classes of aggression-promoting neurons during courtship and aggression, including P1a cells, an intensively studied population of male-specific interneurons. Using HI-FISH in combination with optogenetic activation of aggression-promoting neurons (opto-HI-FISH) we identify candidate downstream functional targets of these cells in a brain-wide, unbiased manner. Finally we compare the activity of P1a neurons during sequential performance of courtship and aggression, using intronic vs. exonic hr38 probes to differentiate newly synthesized nuclear transcripts from cytoplasmic transcripts synthesized at an earlier time. These data provide evidence suggesting that different subsets of P1a neurons may be active during courtship vs. aggression. HI-FISH and associated methods may help to fill an important lacuna in the armamentarium of tools for neural circuit analysis in Drosophila.

Introduction

Monitoring neuronal activity during specific actions or activities is critical to understanding how the brain controls behavior. Measurements within a local brain region of neural activity in non-transparent awake, behaving animals such as mice or fruit flies can be achieved with electrophysiological and calcium imaging techniques (Buzsáki, 2004; Grewe and Helmchen, 2009). In Drosophila, two-photon calcium imaging (TPI) has been widely used (Seelig et al., 2010) because single-unit electrical recording in the fly central brain remains exceedingly difficult (Turner et al., 2008; Wilson et al., 2004). But TPI in Drosophila requires head-fixation, which limits the repertoire of behaviors that flies can perform under these conditions. For example, it is difficult to use TPI to image neuronal activity during the consummatory phases of complex social behaviors, such as male-male aggression or male-female courtship (but see Clowney et al., 2015). Recently, a new method called Flyception2 was introduced for brain imaging in freely walking flies (Grover et al., 2020, 2016). However the spatial resolution of this method is limited, it is invasive, and monitoring the activity of single neurons in freely moving animals is still challenging.

In mice, immediate-early genes (IEGs) such as c-fos, Arc, or Egr1 have been used to map the activity of neurons by visualizing their expression histologically using immunohistochemistry (IHC) or in situ hybridization (ISH) (Morgan and Curran, 1991). More recently, this method has been combined with tissue-clearing and light-sheet microscopy to perform systematic and unbiased whole-brain mapping of neurons activated during different behaviors (Kim et al., 2015; Renier et al., 2016). An extension of this approach, called cellular compartment analysis of temporal activity by fluorescence in situ hybridization (catFISH), enables within-animal comparisons of neuronal populations activated during two different sequential behaviors (Guzowski et al., 1999; Lin et al., 2011).

In Drosophila, by contrast, the application of IEGs to map whole-brain neuronal activation patterns has been limited. In part this is because c-fos is not a particularly good activity marker in flies (Chen et al., 2016). More recently, several studies using fly IEGs such as Hr38 and stripe/Egr1 have been reported (Chen et al., 2016; Fujita et al., 2013; Takayanagi-Kiya et al., 2023; Takayanagi-Kiya and Kiya, 2019). Nevertheless, this technique has been relatively under-utilitized in flies. This may be the case, at least in part, because application of FISH to the adult Drosophila brain has been technically challenging, due to limited probe penetration and high background caused by traditional amplification methods such as Tyramide (Raap et al., 1995).

To circumvent these problems, we have applied in this study an in situ hybridization amplification technique called the Hybridization Chain Reaction (HCR), v.3.0 (Choi et al., 2018) to visualize the expression of Hr38, an IEG expressed by active neurons in the adult male fly brain (Fujita et al., 2013). The design of HCR, which requires conjoint hybridization to target sequences of adjacent pairs of “half-probes” to achieve amplification, affords both greater sensitivity and specificity, allowing detection of low-abundance transcripts with minimal background noise and off-target hybridization. We call this approach HCR3.0-amplified IEG Fluorescent In Situ Hybridization, or “HI-FISH” (Figure 1A). Although we use Hr38 as proof-of-concept, this approach is in principle applicable to any IEG.

Mapping neuronal activities using Hr38 FISH.

(A) Illustrated summary of this study for mapping Hr38 expression induced by social behavior or artificial activation of specific neuronal populations. (B) Representative images of Hr38 expression before (fru-No PS) and after photo-stimulation of fru-GAL4>Chrimson neurons (fru-PS). BDPG4uw-PS represents the images after Photo-stimulation of empty-GAL4>Chrimson neurons. magenta: Chrimson::tdT (native fluorescence), green: Hr38 HCR signals. Scale bar: 100μm. (C) Magnified confocal section images of Hr38 expression (left three columns) from the boxed region (far right column) after photo-stimulation of fru-GAL4>Chrimson neurons. SMP, superior medial protocerebrum; LH, lateral horn; KC, Kenyon cells. Scale bar: 20µm (left three columns) or 100µm (far right column).

Here we demonstrate three different applications of HI-FISH relevant to understanding the neural circuitry underlying social behaviors in flies. First, we have asked whether specific neuronal populations discovered in functional screens for aggression-promoting neurons (Asahina et al., 2014; Hoopfer et al., 2015; Watanabe et al., 2017) are indeed active during natural aggressive behavior, courtship or both. We have further investigated whether such activity is only observed during the contact-mediated consummatory phase of these behaviors, or can also be detected in freely moving flies in the absence of fighting or mating. Second, we have performed unbiased, whole-brain functional mapping of the downstream targets activated by optogenetic stimulation of a neuronal population of interest, a variant method we call “opto-HI-FISH.”. We have applied this method to P1a neurons (Anderson, 2016; Clowney et al., 2015; Hoopfer et al., 2015; Inagaki et al., 2014) a subset of the male-specific P1 class of interneurons originally identified based on their ability to trigger male courtship behavior when stimulated (von Philipsborn et al., 2011). Finally, we have developed a variant of the catFISH technique (Guzowski et al., 1999), using dual-color FISH combining intronic and exonic Hr38 probes, to compare neuronal populations activated during two different sequential behaviors, a method we call “HI-catFISH.” Since experimental activation of P1a neurons can promote both aggression and courtship in pairs of male flies (Hoopfer et al., 2015), we have used HI-catFISH to investigate whether separate or overlapping subpopulations of P1a neurons are active during natural occurrences of these two social behaviors. Together, the results of these studies illustrate the utility of HI-FISH, opto-HI-FISH and HI-catFISH to detect the activation of both known and previously unidentified neurons during naturalistic social behaviors that are not amenable to head-fixation. The results have also yielded new biological insights into the neural circuit-level control of aggression and courtship.

Results

Brain-wide expression of Hr38 detected by HI-FISH

The HCR3.0 method provides greater sensitivity, specificity and lower background than traditional FISH using Tyramide signal amplification (TSA). Application of older FISH methods in the Drosophila adult whole brain has been limited, due to low sensitivity and high background (Wilkie and Davis, 1998). For example, the detection by FISH of low-abundance transcripts such as those encoding GPCRs has been virtually impossible in adult flies up to now. Application of the HCR-FISH method, by contrast, was able to detect mRNA encoding an octopamine receptor Oct-TyrR (Arakawa et al., 1990; Saudou et al., 1990), as well as Dh44, a high-abundance neuropeptide-encoding transcript (Figure 1-figure supplement A and B) (Choi et al., 2018). Background in controls lacking the HCR initiator probes was essentially undetectable.

Next, we investigated the ability of HCR to detect brain-wide expression of Hr38 following artificial (optogenetic) activation of fruitless-GAL4 (Stockinger et al., 2005) neurons expressing the red-shifted opsin Chrimson::tdT. Fru-GAL4 labels ∼2,000 neurons in the adult male brain, providing a large number of potentially activatable cells. Following photostimulation Hr38 signals were detected in virtually all Chrimson::tdT+ neurons as well as in other tdT- neurons (Figure 1B, C and Figure 1-figure supplement C). The latter may be indirectly activated via synaptic inputs from Fru-GAL4+ neurons. Control flies of the same genotype but without photostimulation, or photostimulated flies with an “empty” GAL4 driver (BDPG4uw; Pfeiffer et al., 2008) displayed much lower levels of Hr38 expression.

While performing these experiments, we observed that simply transferring flies to a new arena induced a significant level of Hr38 expression throughout the brain (Figure 1-figure supplement D and E), likely reflecting the stress caused by the manipulation and new environment. Based on this observation, we conducted subsequent experiments after flies were habituated to the testing arena before starting the assays (see Methods). This minor modification proved essential to distinguishing behavior-specific HI-FISH signals from the widespread Hr38 induction caused by transfer of the flies from their home vial to an experimental apparatus.

Mapping of neurons activated by social behaviors

As an initial proof-of-concept application of HI-FISH, we asked whether neuronal subsets initially identified in functional screens for aggression-promoting neurons (Asahina et al., 2014; Hoopfer et al., 2015; Watanabe et al., 2017) were actually active during natural aggressive behavior. These included P1a (Hoopfer et al., 2015), Tachykinin-FruM+ (TkFruM) (Asahina et al., 2014) and aSP2 neurons (Watanabe et al., 2017). Despite the fact that activity in these neuron subsets was shown to be essential for normal aggression, it was formally possible that the behavioral phenotypes caused by their artificial stimulation were due to their supra-physiological activation. Therefore, we performed HI-FISH on fly lines carrying a myr:GFP reporter in the cell populations of interest in males paired with different types of target flies under different conditions. Thirty to 45 minutes after a 20 min interaction with the target fly, we fixed the brains and performed double-label HI-FISH together with fluorescent antibody staining of myr::GFP-expressing neurons (Figure 1A; see Methods).

In the control condition, where the experimental flies were kept alone during the observation period, there was little or no detectable Hr38 expression in P1a, Tk, or aSP2 neurons (Figure 2A-D, “Control”). In contrast, we observed clear Hr38 signals in GFP+ P1a, Tk, and aSP2 neurons after a male-male interaction that included fighting behavior (Aggression: Figure 2A-D, Figure 2-figure supplement. We also observed P1a-negative Hr38+ neurons neighboring P1a neurons during aggression (Figure 2, P1a Hr38). These cells may correspond to other subpopulations of P1 neurons, or to Dsx-expressing pC1 neurons that have also been implicated in aggression (Koganezawa et al., 2016). Note that images are single confocal sections provided for qualitative, illustrative purposes only and should not be interpreted quantitatively. Our conclusions are based on statistically analyzed quantitative data presented in bar and box plots). In the case of courtship with a female, comparably strong Hr38 induction was observed in P1a and aSP2 neurons, while there was no Hr38 expression in Tk neurons (Figure 2A-D, Figure 2-figure supplement; Courtship). These data are consistent with prior observations indicating that the functional manipulation of P1a or aSP2 neurons in the male flies affects both male-male aggression and male-female courtship behavior (Hoopfer et al., 2015; Takayanagi-Kiya and Kiya, 2019; Watanabe et al., 2017), while the manipulation of Tk neurons affects only aggression (Asahina et al., 2014).

Investigation of neurons activated by social behavior with Hr38 FISH.

(A) Hr38 expression in the neurons labeled with P1a-split GAL4, Tk-GAL4, and aSP2-split GAL4 drivers. magenta: myr::GFP, yellow: Hr38 HCR signals. Dotted lines depicted the outlines of myr::GFP labeled cell bodies. Arrows: cell bodies of Hr38 positive neurons. Scale bar: 10µm. (B) (D) Upper, scatter plots indicate expression level (relative intensity) of Hr38 HCR signals; lower bar plots indicate percentage of Hr38 positive neurons (“positive” defined as neurons with a relative signal intensity > 3σ above the average signal intensity of the control condition) in P1a (B), Tk (C), and aSP2 cells (D) after different behavioral episodes. Aggression: aggressive behavior against another male, Dead: interaction with a dead male, No Food: interaction with another male without food (no fighting), Mesh: interaction with another male, separated with mesh (no physical contact), Courtship: courtship behavior with a virgin female. Bar plots indicate the percentage of Hr38+ neurons among GFP-labeled cells. Bars with the same letter are not statistically significantly different; bars with no common letter are significantly different (p < 0.05, chi-square test). Number of neurons and individuals analyzed; P1a-Control: 135 (N = 8), P1a-Aggression: 84 (N = 5), P1a-Dead Male: 126 (N = 7), P1a-No Food: 115 (N = 6), P1a-Mesh: 87 (N = 6), P1a-Courtship: 114 (N = 6), Tk-Control: 45 (N = 4), Tk-Aggression: 69 (N = 6), Tk-Dead Male: 61 (N = 6), Tk-No Food: 63 (N = 6), Tk-Mesh: 82 (N = 8), Tk-Courtship: 66 (N = 6), aSP2-Control: 20 (N = 3), aSP2-Aggression: 36 (N = 4), aSP2-Dead Male: 23 (N = 3), aSP2-No Food: 49 (N = 4), aSP2-Mesh: 36 (N = 4), aSP2-Courtship: 24 (N = 3).

Social interactions in flies consist of multiple phases: approach, investigation, initial contact and progression to consummatory behaviors (Chiu et al., 2021; Yamamoto and Koganezawa, 2013). While electrophysiological recording or calcium imaging can identify neuronal activation that is time-locked to specific behavioral actions, the detection of Hr38 (or any IEG) induction may reflect activation at any or all of these sequential phases due to the long-lived, accumulated expression of the IEG mRNA. To identify the phase(s) of social behavior wherein Hr38 induction occurred, we separated the signals or interactions characteristic of these sequential phases in space rather than time. First, we asked whether the activation of these neurons during a male-male social interaction requires actual fighting behavior, or simply close contact with the target male. Initially, we paired the experimental male fly with a dead male, as a source of chemosensory cues. Because some male-derived cues (e.g., visual motion) may require a live conspecific, we also placed the experimental and tester males in an arena with an agarose substrate. Under these conditions, physical contact but not fighting occurs (Lim et al., 2014). In both “touching but no fighting” conditions, the fraction of Hr38+ cells among P1a, Tk and aSP2 neurons was slightly lower but not statistically significantly different from that measured during actual aggression, while the intensity of Hr38 mRNA expression in positive cells was slightly lower in the No Food condition (Dead Male and No Food: Figure 2B-D, Figure 2-figure supplement). Thus the activation of P1a, Tk, or aSP2 neurons does not require consummatory aggressive actions such as lunging, tussling or boxing (Chen et al., 2002).

Interacting male flies can, in principle, exchange visual, auditory, mechanosensory or chemosensory cues. In the case of chemosensory signals, inter-male aggression has been shown to require both detection of volatile cues by the olfactory system, and of non-volatile cues by the gustatory system (Fernández and Kravitz, 2014; Wang et al., 2011; Wang and Anderson, 2010). While both types of cues are present when an experimental fly is exposed to an intact or a dead male fly, the detection of non-volatile cues requires contact chemosensation (Vosshall and Stocker, 2007). We therefore exposed the experimental males to a target fly separated by a mesh filter which prevented physical contact between the freely moving flies. Interestingly, Hr38 induction in all 3 cell populations under this condition was significantly weaker that that measured during free interaction with a male on agarose, where contact but no fighting occurred (Mesh: Figure 2B-D, Figure 2-figure supplement). Taken together with the relatively strong Hr38 induction observed following contact with a dead male (which presumably provide a source of both olfactory and gustatory pheromones), these data suggest that strong activation of P1a, Tk, and aSP2 neurons requires the integration of both volatile and non-volatile male-derived chemosensory cues. However the results do not exclude the possibility that these neurons exhibit additional activity during contact-mediated aggression, but which is not statistically detectable by this method in comparison to “no fighting” controls.

Detection of Hr38+ cells induced by optogenetic stimulation of P1a neurons in the male brain

Optogenetic and thermogenetic stimulation experiments have shown that P1a interneurons can promote both male-directed aggression and male- or female-directed courtship (Hoopfer et al., 2015; von Philipsborn et al., 2011). While the neural circuitry downstream of P1 neurons that promotes female-directed courtship behavior has been well-characterized (von Philipsborn et al., 2011), the pathway by which P1a neurons can promote aggression is poorly understood. As a first step to identify systematically P1a functional downstream targets, we optogenetically activated P1a neurons using Chrimson:tdT and mapped the brain-wide pattern of Hr38 induction in males (Figure 3A-D), a method we call “opto-HI-FISH (Figure 1A).” To analyze these data, images of Hr38 expression from multiple individuals were registered to a reference brain using antibody labeling of Bruchpilot (Brp), a reference neuropil marker, to generate an average induction pattern (Figure 3E) (Cachero et al., 2010). We then created a voxel-based heat map for the Hr38 expression from the average image. The heat map indicated that neurons around the posterior dorsal region of the central brain were strongly activated by P1a stimulation (Figure 3F).

Investigation of downstream neurons activated by optogenetic activation of P1a neurons.

(A) Illustration of the experimental set up. (B) Expression pattern of Chrimson::tdT driven by P1a split GAL4. (C) Scheme of the data analysis procedure. (D) Time-course study of Hr38 HCR expression after P1a photo-stimulation in P1a labeled neurons and the surrounding area. By 30-60 mins, most of the P1a labeled neurons expressed Hr38 (96.8% of P1a labeled neurons at 30 min). Scale bars: 10µm. (E) Average image of Hr38 expression of control (BDPG4Uw) and P1a-activated brains. 5 individual brains from each conditions were registered into a template brain and averaged. Green: Hr38 HCR signals, Magenta: Brp. Scale bar: 100µm. (F) Heat map analysis of Hr38 expression. Areas depicted by dotted lines represent the areas of PAM neurons (PAM), mushroom bodies (MB), and pC neurons (pC). Scale bar: 100µm. (G) Confocal images of Hr38 expressions in pCd neurons (R41A01), Kenyon cells (OK107), and PAM neurons (0273 or Ddc) of control and P1a activated brains. Green: myr::GFP, Magenta: Hr38. Dotted lines depicted the outlines of cell bodies (pCd) or cluster of neurons (KC, PAM). Scale bar: 10µm.

To validate this approach, we first asked whether we could detect Hr38 induction in pCd neurons, which were previously shown by calcium imaging to be (indirect) targets of P1a neurons (Jung et al., 2020). Double-labeling in pCd-GAL4; UAS-GFP flies using HI-FISH and immunostaining for GFP (see Methods) confirmed that Hr38 was strongly induced in pCd neurons by P1a photostimulation (Figure 3G, pCd). Next, we asked whether we could identify any of the specific cellular targets of P1a neurons in brain regions that exhibited strong induction of Hr38. These regions included Kenyon cells (KCs) and PAM dopaminergic neurons in the mushroom body (MB; Figure 3F). We used the GAL4 drivers OK107 (Connolly et al., 1996) to label KCs, and 0273 and Ddc-GAL4 to label PAM neurons. Following P1a optogenetic activation, brains were processed for double-labeling with HI-FISH and anti-GFP antibody staining to identify the cells of interest. We observed evidence of Hr38 induction in all three cell populations, although the signal was difficult to quantify due to the high density of GFP+ cells (Figure 3G).

To confirm the results of the opto-HI-FISH experiments, we next asked whether PAM neurons and/or Kenyon cells were activated in response to P1a stimulation using in vivo calcium imaging in living, head-fixed flies. We expressed a calcium indicator, jGCaMP7b (Dana et al., 2019) in either Kenyon cells or PAM neurons using the GAL4/UAS system and Chrimson::tdT in P1a neurons using the LexA/LexAop system (Brand and Perrimon, 1993; Lai and Lee, 2006). The neurites of both KCs and PAM neurons showed significant time-locked responses to optogenetic activation of P1a neurons (Figure3-figure supplement 3A-F). We also imaged jGCaMP7b signals from cell bodies of PAM neurons while photo-stimulating P1a neurons, and observed clear responses in both cases as well (Figure 3-figure supplement 3G-J). Although we could not identify the specific activated cell populations among the KCs (data not shown), P1a photo-stimulation evoked a significant rise in ΔF/F in ∼50 % of PAM neurons (Figure 3-figure supplement 3I). These data validate the results from brain-wide Hr38 induction mapping and provide proof-of-concept that brain regions containing functional targets of a cell population of interest (here P1a neurons), as well as specific neuron subtype targets within these regions (KCs and PAM neurons), can be identified by opto-HI-FISH. These results are also consistent with a recent report showing that subsets of PAM dopaminergic neurons receive excitatory inputs from P1 neurons, and assign to KCs a positive valence associated with P1 activation (Shen et al., 2023)

HI-catFISH reveals distinct fighting- and mating-activated P1a neuron subsets

Previous reports showed that artificial activation of P1a neurons promoted both aggression and courtship. The foregoing data indicated that Hr38 was induced in P1a neurons after both natural aggressive and courtship behavior (Figure 2, Figure 2-figure supplement). An important outstanding question is whether the aggression-and courtship-activated P1a populations are the same or different (Anderson, 2016). Because individual P1 neuron subtypes cannot be distinguished by FISH, between-animal comparisons of Hr38 induction following fighting or mating cannot distinguish these two possibilities. To compare directly which P1a neurons are active during naturalistic mating vs. aggression, we developed an Hr38 analog of catFISH (HI-catFISH) to compare neurons activated during aggression and courtship in the same animal (Figure 4A). The catFISH method exploits the different kinetics of nuclear-cytoplasmic transport and of IEG mRNA stability (Guzowski et al., 1999) to distinguish neurons activated during a behavior performed just prior to analysis (nuclear pre-mRNA), from those activated during a different behavior performed 30 min before the second (cytoplasmic mRNA); cells activated during both behaviors contain both cytoplasmic and nuclear signals (Figure 4B). In mice, we have used catFISH with c-fos to compare hypothalamic neurons activated during aggression vs. mating in males (Lin et al., 2011).

Drosophila catFISH reveals P1a subpopulations activated by two different social behaviors.

(A) Experimental design of the fly catFISH. Each fly experienced the 1st behavior (aggression or courtship) followed by 2nd behavior (aggression or courtship), and the HCR was performed with the Hr38 probes targeted for the exon or the intron sequence. (B) Time-course of Hr38 signals detected by the exon- or the intron-targeted Hr38 HCR probes and expected Hr38 signals in each neurons. (Top) The intron-targeted Hr38 HCR signals (magenta) induced by 1st behavior disappear at the timing of HCR analysis while the exon-targeted Hr38 signals (blue) remain in the cytoplasmic area. (Middle) Both intron-targeted and exon-targeted signals are present in the nucleus. (Bottom) Both intron-targeted and exon-targeted signals induced by the 2nd behavior are present in the nucleus, and exon-targeted Hr38 signals induced by the 1st behavior are present in the cytoplasmic area. In ‘Neurons active during both’, the exon probe signals from the 1st behavior decline (dotted blue line *1) after 60 mins (also see Fig. 4B-supplement). However, it is canceled out as the signals from 2nd behavior are increased (dotted blue line *2). As a result, the combined Hr38 exon probe signals (solid blue line) remain high at the time point for Hr38 HCR. (C) Representative images of Hr38 HCR signals in P1a neurons after six different behavioral conditions. Green: myr::GFP, Magenta: intron-targeted Hr38 HCR signals: exon-targeted Hr38 signals. Scale bar: 10µm. The images shown are for illustration purposes only and represent single optical sections covering only a portion of the P1a neuronal cluster. They exclude P1a cells present in other Z-planes. Quantification (D) was based on signals measured in all P1a cells contained within an entire Z-stack. See also Figure 4 - figure supplement D. (D) Fraction of responsive P1a neurons across indicated conditions. N.R.: not responded, Nuc: nuclear signal only (responded to only second target), Cyto: cytoplasmic signal only (responded to only first target), Both: nuclear and cytoplasmic signal (responded to both first and second target).

We detected neuronal populations activated during two consecutive social behaviors using HI-FISH containing Hr38 intronic and exonic probes amplified using different fluorophores. In two color FISH, unspliced pre-mRNAs were detected as nuclear dots labeled by both probes, while mRNA was detected as a more diffuse cytoplasmic signal detected by only the exonic probe (Figure 4B). A time course study using opto-HI-FISH with these two different Hr38 probes indicated that the primary RNA transcripts first appeared in P1a nuclei ≤15 min after optogenetic activation, and disappeared by 60 min. Conversely, cytoplasmic processed mRNAs started to accumulate in the cytoplasm 30 min after activation and disappeared by 2-3 hours after stimulation (Figure 4-figure supplement A-C).

Having established this calibration of the temporal resolution of HI-catFISH, we compared labeling during free social interactions with a male or a female target. Following pre-habituation (see above), a naïve experimental male fly was placed into a behavioral arena, and a male or female fly was introduced as a first target. After 20 minutes of free social interactions the target fly was removed from the arena and, after a further 30 min recovery period, replaced by a second target fly (male or female; Methods). All four combinations of sex of the 1st and 2nd target fly, and two additional control conditions, in which the experimental flies were exposed to only the 1st target fly (Male -> None, Female -> None), were tested (Fig. 4C). For the two sequential behaviors condition, ten minutes after exposure to the second target fly (and ∼1 hr after exposure to the first target fly), the experimental fly was sacrificed and analyzed by dual-color HI-FISH Hr38 (Figure 4A and B).

As expected, many of the neurons activated by the first target were re-activated by the second target (nuclear and cytoplasmic signal), when the sex of the two target flies was the same (Figure 4C, Male -> Male, Female -> Female, arrowheads and Figure 4D, “Both” vs. “Cyto”). In contrast, few neurons exhibited nuclear Hr38 transcripts detected with the intronic probe when the experimental flies were exposed to the only 1st target fly but sacrificed after an additional 60 minutes (Figure 4C, Male -> None, Female -> None, “Nuc” or “Both”). By contrast, many neurons activated by the first target fly (cytoplasmic signal) were not activated by the second target fly (nuclear signal), when the target sexes were different (Figure 4C, Female -> Male, Male -> Female, arrows and Figure 4D, “Cyto” vs. “Both”). Responses to males were clearly detected in a subset of P1a neurons whether the male was presented first (Figure 4C, Male→Female; GFP+; dashed oval outlines containing blue diffuse cytoplasmic signal) or second (Figure 4C, Female→Male; magenta nuclear dots). In addition, there were male-activated cells nearby that were not labeled by the P1a-split GAL4 driver. In general, these male HI-FISH signals did not overlap with the HI-FISH signals evoked by female targets. (We note, however, that relatively few neurons were activated by the female fly when the sex of the two targets was different, regardless of whether the female was presented first or second; the reasons for this are currently unclear but may reflect a suppressive effect of aggression neuron activation on courtship neuron activity, as suggested by earlier studies (Hoopfer et al., 2015).) These data suggest that there may be distinct P1a subpopulations that are activated during either male-male aggression or male-female courtship.

Discussion

Drosophila is a powerful model system for identifying and characterizing neural circuits that control innate behaviors. With recent progress in two-photon imaging techniques using genetically encoded calcium sensors, it has become possible to monitor activity from genetically labeled neurons at single-cell resolution in head-fixed flies (Seelig et al., 2010). Although single-neuron imaging in unrestrained behaving animals has been achieved in mammalian studies (Flusberg et al., 2008; Remedios et al., 2017; Ziv et al., 2013) most imaging studies with single-cell resolution in Drosophila require head-fixed conditions. Therefore, the repertoire of naturalistic behaviors that can studied using this approach – in particular social behaviors dependent on chemosensory cues (but see Hindmarsh Sten et al., 2021)– remains limited.

In mammalian systems, the discovery that IEGs such as c-fos are induced by neuronal activity (Greenberg et al., 1985; Greenberg and Ziff, 1984) has enabled brain-wide functional mapping of neuronal activation following specific behaviors (Morgan and Curran, 1991; Renier et al., 2016). This method and its derivatives have had a transformative impact on the identification of behaviorally relevant neural circuits in rodents. In Drosophila, by contrast, few if any studies describe the application of IEGs for brain-wide activity mapping. In this report, we applied the HCR3.0 RNA-FISH technique (Choi et al., 2018) to visualize the expression of a fly IEG, Hr38 in the entire adult male brain following different manipulations (see also Shao et al., 2019). We present data providing proof-of-concept that HI-FISH can be used to detect the activation by naturalistic stimuli of neurons previously identified in thermogenetic behavioral activation screens, as well as to identify candidate downstream targets of these neurons following their optogenetic activation (opto-HI-FISH). We also compared neuronal subsets within a small identified population (P1a neurons) that is activated during two different social behaviors, using HI-catFISH.

Monitoring Neuronal Activities during Social Behaviors

Previous studies reported that male-specific P1a, Tk, and aSP2 neurons promote male-male aggression (Asahina et al., 2014; Hoopfer et al., 2015; Watanabe et al., 2017). In addition, P1a and aSP2 neurons promote male-female courtship (Hoopfer et al., 2015; Takayanagi-Kiya and Kiya, 2019). Artificial manipulation of the activity of these neurons affected social behaviors and demonstrated their causal roles in these behaviors. However, there was no direct evidence showing that these neurons are normally activated during the specific behaviors they can promote.

Our HI-FISH data demonstrated that P1a and aSP2 neurons are active during male-male aggression and male-female courtship, consistent with the behavioral effects of activating these neurons (Figure 2) (Hoopfer et al., 2015; Watanabe et al., 2017). In contrast, TkFruM neurons were only active during male-male aggression (Figure 2), consistent with their lack of a functional role in courtship (Asahina et al., 2014). Furthermore, we show that P1a, Tk, and aSP2 neurons can apparently be activated by chemosensory cues from an opponent male fly in the absence of attack behavior (Figure 2). These data suggest that P1a, Tk, and aSP2 neurons may serve to process and integrate sensory cues from the opponent and relay that information to the downstream neurons that encode the consummatory phase of aggression (Chiu et al., 2021). However they do not exclude that these neurons may play a role during overt attack.

Although the time resolution of IEG-based activity monitoring falls far short of that of calcium imaging, our data show that one can nevertheless use HI-FISH to separate activation that occurs during the appetitive vs. consummatory phases of male-male aggression, by judicious manipulation of experimental conditions.

Identification of downstream target neurons of genetically labeled neurons by opto-HI-FISH

Artificial activation of P1a neurons induces male-female courtship and an internal state that promotes male-male aggression (Bath et al., 2014; Hoopfer et al., 2015; Inagaki et al., 2014). Furthermore, artificial activation of male specific P1 neurons mimics courtship-induced place preference as well as supporting appetitive olfactory conditioning (Shen et al., 2023). It is critical to identify neurons that are functionally downstream of P1a cells to understand how these neurons can promote this state. In this study, using opto-HI-FISH we identified pCd neurons, Kenyon cells and PAM neurons as downstream targets of P1a neurons. These results were confirmed by two-photon calcium imaging analysis (Figure 3, Figure 3-figure supplement). Importantly, recent studies have reached the same conclusions regarding neuronal populations using different approaches (Jung et al., 2020; Shen et al., 2023). These findings provide additional support for our methodology. The heat map illustrates the presence of additional potential P1a -follower populations beyond the three interrogated here (Figure 3F). Further analysis will unveil the detailed characteristics of these potential downstream neurons of P1a neurons. An important caveat is that not all active Drosophila neurons may express Hr38. Act-seq data from mice indicates that different neuron subsets activated during social behaviors express different IEGs to different extents (Kim et al., 2019).

Functional heterogeneity of P1a populations during Social Behaviors

In addition to identifying functional downstream targets of P1a neurons, characterizing cellular diversity among P1a neurons and how this functional heterogeneity maps onto their connectivity with different downstream targets is critical to understanding how these cells can induce both courtship and aggressive behavior. Optogenetic activation of P1a neurons at a low frequency promotes aggression towards males, while high-frequency stimulation promotes male-directed courtship during stimulation, followed by aggression in the inter-stimulation intervals (Hoopfer et al., 2015). P1a neurons may control these two behaviors according to their activity level and/or the subtype of neuron activated in the population. Alternatively, P1a neurons may indirectly promote fighting by inhibiting aggression-promoting neurons during courtship (Koganezawa et al., 2016); post-inhibitory rebound in such neurons following the offset of P1a optogenetic stimulation may trigger aggressive behavior (Anderson, 2016).

Our preliminary Hr38 catFISH results suggest the presence of male- or female-responsive subsets within the P1a population, which consists of 8-10 neurons/hemibrain (Figure 4). To our knowledge, this is the first evidence that P1a neurons are activated during male-male social interactions. Previous studies using calcium imaging of head-fixed flies suggested that P1a neurons were inhibited by the male-specific pheromone, 11-cis-vaccenyl acetate (cVA; Clowney et al., 2015). The reason for this apparent discrepancy is not yet clear. The molecular identity of the male- vs. female-responsive P1a subsets remains to be determined. It has been suggested that FruM+ P1 neurons exclusively control courtship, while Dsx+ P1 neurons (also called pC1 neurons; Koganezawa et al., 2016)) control aggression. However triple-intersection experiments have demonstrated that thermogenetic (Hoopfer et al., 2015) or optogenetic (Watanabe et al., 2017) activation of FruM+ P1a neurons, which comprise only 1-4 neurons/hemibrain, can also promote aggression. Therefore, while our results support the idea that different P1 subsets control aggression vs. courtship, they do not support the view that FruM+ P1 neurons only control the latter and not the former social behavior.

While this paper was in the final stages of preparation, a manuscript by Takayanagi-Kiya et al. appeared that described a method for labeling neurons activated during innate behaviors using a different immediate early gene, stripe/egr-1 (Takayanagi-Kiya et al., 2023). To analyze neuronal activities in different behavioral contexts, that study utilized a GAL4 driver for the stripe/egr-1 gene to label neurons activated by the first behavior and immunostaining for stripe protein to label the second behavior. The flies were exposed to the first target for 24 hours, followed by exposure to the second target for 2 hours. In contrast, our approach for HI-catFISH is based purely on detecting the expression of the IEG mRNA using a state-of-the-art amplification method, HCR v3.0 (Figure 1-figure supplement B) (Choi et al., 2018). This enables us to monitor neuronal activity during two different behaviors with much higher temporal resolution than the analogous method of Takayanagi-Kiya. Furthermore, by combining our HI-catFISH with their protein expression-based method, it is theoretically possible to monitor neurons activated by four different behavioral contexts. Thus, the two methods complement each other.

In mice, distinct subpopulations of estrogen receptor 1 (Esr1)-expressing neurons in the ventrolateral subdivision of the ventromedial hypothalamus (VMHvl) are activated during male-directed aggression versus female-directed mating (Karigo et al., 2021; Remedios et al., 2017). Our observation that apparently distinct subpopulations of P1a neurons are activated during aggression versus courtship in Drosophila strengthens the analogy between these two control nodes for social behavior (Anderson, 2016). Application to P1a neurons of Act-Seq, a method that uses scRNAseq to identify transcriptomic cell types that are active during different behaviors as determined by IEG expression (Kim et al., 2019; Wu et al., 2017), may help to identify transcriptomic and functional diversity within the P1a population. Such correlative studies may be extended to functional perturbations by using transcriptomic data to identify more specific intersectional GAL4 drivers for distinct P1a subsets. Together, our studies illustrate how HI-FISH can provide a systematic, unbiased and brain-wide approach to identifying entry points to circuits controlling specific social behaviors, which can be integrated eventually with imaging, connectomic, transcriptomic and functional approaches to circuit analysis.

Materials and methods

Fly stocks

A list of the full genotypes for the flies used in this research is in Supplementary file 1.

R15A01-iLexA (improved LexA) was generated for this study. Briefly, the R15A01 enhancer fragment (Jenett et al., 2012) was cloned into pBPnlsLexA::p65::GADUw (Chiu et al., 2021).

20xUAS-IVS-Syn21-Chrimson::tdTomato3.1 (attP2) and 13xLexAop2-IVS-Syn21-Chrimson::tdTomato3.1 (su(Hw)attP5) were obtained from G. Rubin (Janelia Research Campus). 0273-GAL4 (Burke et al., 2012) was obtained from S. Waddell (The University of Oxford). fruGAL4 (Stockinger et al., 2005) was from B. Dickson (Janelia Research Campus).

The following strains were obtained from Bloomington Drosophila Stock Center (Indiana University); R15A01-AD (attP40), R71G01-DBD (attP2), R41A01-GAL4 (attP2), OK107, Ddc-GAL4, 20xUAS-IVS-jGCaMP7b (VK00005).

Optogenetic activation in behaving flies

A detailed description of the setup used for optogenetic activation was described previously (Inagaki et al., 2014; Watanabe et al., 2017). Flies were raised at 25℃ and collected on the day of eclosion and kept in the dark for 7-10 days on the fly media containing 0.4 mM all trans-Retinal (Merck, St. Louis, MO). The experiments were performed with the “8-well” chamber (16 mm diameter x 10 mm height) with IR backlight. (855 nm, SOBL-6x4-850, SmartVision Lights, Norton Shores, MI). A 655 nm 10 mm Square LED (Luxeon Star LEDs Quadica Developments, Brantford, Canada) was used for photostimulation. The stimulation protocols for each experiment are described in the figures or figure legends.

Behavioral Assays

Flies were maintained at 9AM:9PM Light:Dark cycle. For male-male aggression assays and male-female courtship assays, the “8-well” acrylic chamber was used as described previously (Inagaki et al., 2014; Watanabe et al., 2017). The walls of the chamber were coated with a fluoropolymer resin (PTFE-30, Insect-a-Slip) (BioQuip Products, Rancho Dominguez, CA). The clear top ceiling plate was coated with Sigmacote (Merck, St. Louis, MO) to prevent flies from walking on the surfaces. The floor was covered with a layer of apple juice/sucrose-agarose food. The arena was illuminated with IR backlighting.

Two-photon calcium imaging

Calcium imaging with optogenetic activation was performed as described (Inagaki et al., 2014; Jung et al., 2020). A deep red (660 nm) fiber-coupled LED (Thorlabs, Newton, NJ) with a band-pass filter (660 nm, Edmund Optics, Barrington, NJ) was used for light source to activate Chrimson.

Flies were collected on the day of eclosion and kept in the dark on the fly media containing 0.4 mM all trans-Retinal (Merck, St. Louis, MO). 7-10 days old flies were cold-anesthetized using a temperature-controlled stage and mounted on a plastic plate with a small window using wax. The cuticle of the dorsal side of the head was removed by forceps. Flies were then placed beneath the objective.

Fly saline (108 mM NaCl, 5 mM KCl, 4 mM NaHCO3, 1 mM NaH2PO4, 5 mM trehalose, 10 mM sucrose, 5 mM HEPES, 0.5 mM CaCl2, 2 mM MgCl2, pH = 7.5) was used to bathe the brain during imaging.

Immunohistochemistry and HCR fluorescence in situ hybridization (FISH)

In situ HCR was performed as previously described with minor modifications (Choi et al., 2018, also see https://www.moleculartechnologies.org/supp/HCRv3_protocol_generic_solution.pdf). For the hybridization buffer, formamide was substituted with an equal volume of 8M urea solution for better preservation of tissue morphology (Sinigaglia et al., 2018).

To process multiple individuals, we collected flies (∼ 8 individuals at once), soaked them in ice-cold ethanol to remove the wax on the cuticle, and opened the head cuticle of each fly with minimal time-lag among samples in the ice-cold fixative (4% paraformaldehyde in PBS). After brief fixation, the brain from each individual was dissected and fixed for 2 hours at 4℃.

For combination of immunohistochemistry (IHC) and HCR FISH, we performed IHC as described previously before HCR FISH procedure in RNase free condition (Watanabe et al., 2017). For antibody incubation, we added RNasin Plus (5% of final volume, Promega, Madison, WI). After immunostaining, the samples were post-fixed with 4% formaldehyde in PBS and proceeded for HCR. HCR probes were purchased from Molecular Technologies (https://www.moleculartechnologies.org/). The identifiers for each probes are as follows; Hr38 (exon): Hr38_MT 2627/B655 (B2 Amplifier) and Hr38 3278/C629 (B5 Amplifier), Hr38 (intron): HR38intron 3736/D801 (B1 Amplifier), Dh44: Dh44 3174/C241 (B3 Amplifier), Oct-TyrR: Oct-TyrR_CDS 3008/B815 (B2 Amplifier). We tested two different Hr38 probes (2627/B655 and 3278/C629) and obtained the same results. Images of HCR3.0 data presented in this manuscript were obtained using the Hr38_MT 2627/B655 probe set. The samples were mounted in 88% (w/v) Histodenz in 0.02 M Phosphate buffer (pH 7.5 with NaOH) with 0.1% Tween-20 and 0.01% sodium azide (Yang et al., 2014).

The antibodies used were as follows: rat anti-DN Cadherin (1:50, DN-EX #8, developed by T. Uemura, obtained from Developmental Studies Hybridoma Bank, University of Iowa, Iowa City, IA), mouse anti-Bruchpilot (1:50, nc82, developed by E. Buchner, obtained from Developmental Studies Hybridoma Bank, University of Iowa, Iowa City, IA), rabbit anti-GFP (1:1,000, A11122, Thermo Fisher Scientific, Waltham, MA), Goat anti-Mouse IgG (H+L) Alexa Fluor 488 (1:1,000, A-11001, Thermo Fisher Scientific, Waltham, MA), Goat anti-Mouse IgG (H+L) Alexa Fluor 568 (1:1,000, A-11004, Thermo Fisher Scientific, Waltham, MA), Goat anti-Mouse IgG (H+L) Alexa Fluor 633 (1:1,000, A-21050, Thermo Fisher Scientific, Waltham, MA), Goat anti-Rabbit IgG (H+L) Alexa Fluor 488 (1:1,000, A-11008, Thermo Fisher Scientific, Waltham, MA), Goat anti-Rabbit IgG (H+L) Alexa Fluor 488 (1:1,000, A-11011, Thermo Fisher Scientific, Waltham, MA), Goat anti-Rabbit IgG (H+L) Alexa Fluor 633 (1:1,000, A-21070, Thermo Fisher Scientific, Waltham, MA)

Confocal serial optical sections were obtained with a Fluoview FV3000 Confocal Microscope (Olympus, Tokyo, Japan).

For the Hr38 exonic probe, we first measured average fluorescence intensity within the cell boundary using Fiji (Schindelin et al., 2012; Schneider et al., 2012) and calculated the relative intensity to the background signals and then defined the positive cells whose signals were above 3σ. For each optical section we defined as the background signal the area outside the region with clearly Hr38+ cells, because the intensity of the background signal varies depending on the depth of the optical section.

For image registration, brain images were registered to T1 template brain (Yu et al., 2010) using the CMTK registration plugin of Fiji (https://flybrain.mrc-lmb.cam.ac.uk/dokuwiki/doku.php?id=warping_manual:registration_gui). To create average images for each condition, MATLAB was used to generate averaged images of each z-plane from different individuals. For voxel-based heat map analysis of the Hr38 expression, the median filter (pixel size = 10) in Fiji was applied to the averaged images. Fiji and Fluorender software (Yong Wan et al., 2009) were used to create z stack images.

catFISH

To prepare target flies, 4 to 6 particles of small iron filings (Delta Education, #060-0313 or equivalent) were attached on the dorsal side of the thorax of 6-8 days old male or female flies using UV-activated glue under CO2 anesthesia one day before the experiment. This allowed us to remove the target fly from a behavioral arena through a small hole using a neodymium magnet.

6-8 days single-housed male flies were used for the experimental flies. A male fly was introduced in the behavioral arena and allowed to habituated to the new environment for ∼ 3 hours. The first target fly was introduced into the arena, and the experimental fly was exposed to the target for 20 mins. After the exposure, the target fly was removed from the arena as describe above. The experimental fly was given a 30-min interval before the second target was introduced. After 20 min exposure, the second target was removed, and Hr38 expression was analyzed 10 min after the target fly removal (Figure 4A and B). Neurons were considered as positive for the Hr38 intronic probe when a clear dot signal with the intronic probe that co-localized with the Hr38 exonic signal was observed.

Statistical Analysis

Statistical analyses were performed using MATLAB (MathWorks, Natick, MA) and Python (Python Software Foundation). The detailed information including the number of samples, the statistical method, and the p value for each experiment is indicated in the figure legends.

Calcium imaging data analysis

T-series images were manually screened and omitted from further analysis in cases where images were shifted in Z-direction, indicative of Z-motion artifacts. For image registration on the X-Y axis, T-series images were opened in Fiji by importing as Image Sequences. The Template Matching Plugin was used for motion correction. Briefly, from the Plugins menu, we selected Template Matching, then selected Align Slices in Stack. We then selected Normalized Correlation Coefficient as a Matching method and left other parameters as default values. We defined a landmark region containing the ROIs by clicking and dragging to draw a rectangle, and initiated the registration process. After image registration, ROIs (region of interest) corresponding to nerve bundle or cell bodies of labeled neurons were manually selected and the fluorescence intensity in the ROIs was measured using Fiji. The average fluorescence intensity during the first 15 s of image acquisition was used as the baseline to calculate the ΔF/F. The mean of ΔF/F during the indicated periods was calculated and visualized using Python. For cell body analysis, we defined the positive response if the average ΔF/F during the photostimulation was statistically higher than the baseline (p < 0.01). The strong response neurons (ΔF/F > 0.4) and the intermediate neurons (0.15 < ΔF/F ≤ 0.4) were defined based on the average ΔF/F value during the stimulation, respectively.

Supporting information